---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

popover_notes <- list(
  workers = list(
    number = 1,
    content = paste0(
      "<code>workers</code> controls how many parallel R processes <code>shapr</code> may use. Configure it through ",
      "<code>future::plan()</code> ",
      "before calling <code>explain()</code>, for example <code>future::plan(\"multisession\", workers = 4)</code>. ",
      "The default is one worker. In a fresh R session, this is equivalent to ",
      "<code>future::plan(\"sequential\")</code> or not calling <code>future::plan()</code>."
    )
  ),
  batching = list(
    number = 2,
    content = paste0(
      "<code>min_n_batches</code> and <code>max_batch_size</code> control the number of batches in each iteration, while ",
      "<code>max_batch_cube_size</code> limits dense per-batch arrays for Gaussian, copula, and empirical. ",
      "Pass the controls through <code>extra_computation_args</code>, ",
      "for example <code>list(min_n_batches = 20)</code>. The defaults are 10, 10, and one million, respectively. ",
      "Progress is reported once per batch, so fewer batches give less frequent updates."
    )
  )
)

popover_note <- function(key, label = key) {
  note <- popover_notes[[key]]
  if (is.null(note)) {
    stop("Unknown popover note: ", key)
  }

  formatted_labels <- paste0("<code>", htmltools::htmlEscape(label), "</code>")
  formatted_label <- if (length(formatted_labels) == 1) {
    formatted_labels
  } else if (length(formatted_labels) == 2) {
    paste(formatted_labels, collapse = " and ")
  } else {
    paste0(paste(utils::head(formatted_labels, -1), collapse = ", "), ", and ", utils::tail(formatted_labels, 1))
  }

  htmltools::HTML(sprintf(
    paste0(
      '%s<sup><a class="footnote-ref" tabindex="0" role="button" ',
      'data-bs-toggle="popover" data-bs-content="%s">[%s]</a></sup>'
    ),
    formatted_label,
    htmltools::htmlEscape(note$content, attribute = TRUE),
    note$number
  ))
}

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, save_explanations := NULL]
benchmark_results[
  is.na(model_variant) | model_variant == "default",
  model_variant := data.table::fifelse(dataset == "numeric", "xgb", "ranger")
]
benchmark_results[, approach_args := sub("variant=", "regression_variant=", approach_args, fixed = TRUE)]
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)

categorical_columns <- c(
  "dataset",
  "backend",
  "model_variant",
  "sweep",
  "approach_args",
  "pair_key",
  "pair_role"
)
benchmark_results[, (categorical_columns) := lapply(.SD, factor), .SDcols = categorical_columns]

batch_limit_columns <- c("max_batch_size", "max_batch_cube_size")
benchmark_results[, (batch_limit_columns) := lapply(.SD, factor), .SDcols = batch_limit_columns]

diagnostic_columns <- c(
  "cpu_user_median",
  "ram_mb_max",
  "bash_wall_median",
  "data_load_median"
)
benchmark_results[, (diagnostic_columns) := NULL]

headline_columns <- c("wall_median", "ram_mb_median")
table_columns <- setdiff(names(benchmark_results), c("n_iterations_median", headline_columns))
table_columns <- append(table_columns, "n_iterations_median", after = match("iterative", table_columns))
data.table::setcolorder(
  benchmark_results,
  c(table_columns, headline_columns)
)

hidden_columns <- c(
  "max_batch_size",
  "max_batch_cube_size",
  "backend",
  "dt_threads",
  "group_size",
  "sweep",
  "pair_key",
  "pair_role",
  "wall_iqr",
  "n_iterations_median"
)
hidden_column_indices <- match(hidden_columns, names(benchmark_results)) - 1

column_descriptions <- c(
  dataset = "Benchmark dataset type.",
  approach = "Conditional-distribution approach passed to explain().",
  n_features = "Number of features in x_train and x_explain.",
  max_n_coalitions = "Maximum number of coalitions passed to explain().",
  n_MC_samples = "Number of Monte Carlo samples passed to explain().",
  min_n_batches = "Minimum batches per iteration, passed through explain(extra_computation_args = list(...)).",
  max_batch_size = "Maximum coalitions per batch, passed through explain(extra_computation_args = list(...)).",
  max_batch_cube_size = "Dense-array size limit, passed through explain(extra_computation_args = list(...)).",
  workers = "Parallel workers configured with future::plan() before calling explain().",
  backend = "future backend used for parallel execution.",
  dt_threads = "Number of data.table threads.",
  n_train = "Number of rows in x_train.",
  n_explain = "Number of rows in x_explain.",
  iterative = "Whether iterative estimation was requested in explain().",
  n_iterations_median = "Median number of Shapley estimation iterations.",
  group = "Whether group-wise explanations were requested in explain().",
  group_size = "Number of consecutive features per generated group.",
  model_variant = "Prediction model being explained: xgb, xgb_large, ranger, or linear.",
  sweep = "Benchmark block that generated the configuration.",
  approach_args = paste(
    "Arguments passed through explain(...); regression_variant selects a registered regression estimator",
    "and tuning recipe."
  ),
  pair_key = "Identifier for a matched iterative/fixed-budget comparison.",
  pair_role = "Role in a matched iterative/fixed-budget comparison.",
  n = "Number of successful measured replicates.",
  wall_median = "Median runtime of explain() across replicates, in seconds.",
  wall_iqr = "Interquartile range of explain() runtimes across replicates, in seconds.",
  ram_mb_median = paste(
    "Median peak RAM used during the full benchmark run, including loaded data,",
    "the fitted prediction model, and workers; 1 MiB = 1.0486 MB."
  )
)
stopifnot(setequal(names(column_descriptions), names(benchmark_results)))
column_descriptions <- column_descriptions[names(benchmark_results)]

display_name_overrides <- c(
  "Iterations" = "n_iterations_median",
  "Runtime (s)" = "wall_median",
  "Runtime IQR (s)" = "wall_iqr",
  "Peak RAM (MiB)" = "ram_mb_median"
)
column_display_names <- stats::setNames(names(benchmark_results), names(benchmark_results))
column_display_names[unname(display_name_overrides)] <- names(display_name_overrides)

explain_columns <- c("approach", "max_n_coalitions", "n_MC_samples", "iterative", "group", "approach_args")
extra_computation_columns <- c("min_n_batches", "max_batch_size", "max_batch_cube_size")
column_guide_names <- stats::setNames(
  sprintf("<code>%s</code>", column_display_names),
  names(column_display_names)
)
column_guide_names[explain_columns] <- sprintf(
  '<a href="../reference/explain.html"><code>%s</code></a>',
  column_display_names[explain_columns]
)
column_guide_names[extra_computation_columns] <- sprintf(
  '<a href="../reference/get_extra_comp_args_default.html"><code>%s</code></a>',
  column_display_names[extra_computation_columns]
)
column_guide <- data.frame(
  Column = unname(column_guide_names),
  Description = unname(column_descriptions),
  check.names = FALSE
)

column_tooltip_map <- paste(
  sprintf(
    "%s: %s",
    encodeString(unname(column_display_names), quote = "\""),
    encodeString(unname(column_descriptions), quote = "\"")
  ),
  collapse = ", "
)

discrete_slider_columns <- c(
  "n_features",
  "max_n_coalitions",
  "n_MC_samples",
  "min_n_batches",
  "workers",
  "dt_threads",
  "n_train",
  "n_explain",
  "group_size",
  "n",
  "n_iterations_median"
)
discrete_slider_map <- paste(
  vapply(discrete_slider_columns, function(column) {
    values <- sort(unique(benchmark_results[[column]]))
    sprintf(
      "%d: [%s]",
      match(column, names(benchmark_results)) - 1,
      paste(values, collapse = ", ")
    )
  }, character(1)),
  collapse = ", "
)
benchmark_table_callback <- DT::JS(paste0(
  "function(settings, json) {",
  "var discreteValues = {", discrete_slider_map, "};",
  "var columnTooltips = {", column_tooltip_map, "};",
  "var table = this.api();",
  "var $container = $(table.table().container());",
  "var $header = $container.find('.dataTables_scrollHead thead').first();",
  "var applyHeaderTooltips = function() {",
  "table.columns().every(function(columnIndex) {",
  "var columnName = table.settings()[0].aoColumns[columnIndex].sTitle;",
  "var tooltip = columnTooltips[columnName];",
  "if (tooltip) $(this.header()).attr('title', tooltip);",
  "});",
  "};",
  "var applyDiscreteSliders = function() {",
  "Object.keys(discreteValues).forEach(function(columnIndex) {",
  "var values = discreteValues[columnIndex];",
  "var visibleIndex = table.column.index('toVisible', Number(columnIndex));",
  "if (visibleIndex === null) return;",
  "var range = {min: values[0], max: values[values.length - 1]};",
  "values.slice(1, -1).forEach(function(value, index) {",
  "var position = 100 * (index + 1) / (values.length - 1);",
  "range[position.toFixed(6) + '%'] = value;",
  "});",
  "var $slider = $header.find('tr').eq(1).children().eq(visibleIndex).find('.noUi-target');",
  "if (!$slider.length || $slider.data('benchmarkDiscrete')) return;",
  "var currentValues = $slider.val().map(Number);",
  "$slider.noUiSlider({",
  "start: currentValues,",
  "range: range,",
  "snap: true",
  "}, true);",
  "$slider.data('benchmarkDiscrete', true);",
  "var $popup = $slider.parent();",
  "$popup.children('.benchmark-discrete-values').remove();",
  "$('<div class=\"benchmark-discrete-values\"></div>')",
  ".text('Available values: ' + values.join(', ')).appendTo($popup);",
  "});",
  "};",
  "var syncFixedFilterCells = function() {",
  "var $rows = $header.find('tr');",
  "var $labels = $rows.eq(0).children();",
  "var $filters = $rows.eq(1).children();",
  "$filters.removeClass('dtfc-fixed-left dtfc-fixed-right').css({left: '', right: '', position: '', zIndex: ''});",
  "$labels.each(function(index) {",
  "var $label = $(this);",
  "var side = $label.hasClass('dtfc-fixed-right') ? 'right' :",
  "  ($label.hasClass('dtfc-fixed-left') ? 'left' : null);",
  "if (!side) return;",
  "var $filter = $filters.eq(index);",
  "$filter.addClass('dtfc-fixed-' + side).css({position: 'sticky', zIndex: 3});",
  "$filter.css(side, $label.css(side));",
  "});",
  "};",
  "document.addEventListener('mousedown', function(event) {",
  "var $input = $(event.target).closest('.dtfc-fixed-right input');",
  "if (!$input.closest('.benchmark-table-wide').length) return;",
  "event.preventDefault();",
  "var $popup = $input.parent().next();",
  "var shouldOpen = !$popup.is(':visible');",
  "$header.find('.dtfc-fixed-right input').each(function() {",
  "var $otherPopup = $(this).parent().next();",
  "if (!$otherPopup.is($popup)) $otherPopup.hide();",
  "});",
  "if (shouldOpen) {",
  "$input.closest('.dataTables_scrollHead').addClass('benchmark-filter-open');",
  "$popup.css({left: 'auto', right: 0}).show();",
  "} else {",
  "$popup.hide();",
  "$input.closest('.dataTables_scrollHead').removeClass('benchmark-filter-open');",
  "}",
  "}, true);",
  "$(document).on('mousedown.benchmarkFilters', function(event) {",
  "if ($(event.target).closest('.dtfc-fixed-right').length) return;",
  "$header.find('.dtfc-fixed-right input').each(function() {",
  "$(this).parent().next().hide();",
  "});",
  "$header.closest('.dataTables_scrollHead').removeClass('benchmark-filter-open');",
  "});",
  "table.on('column-sizing.dt.benchmarkFilters column-visibility.dt.benchmarkFilters', function() {",
  "window.setTimeout(function() {",
  "applyHeaderTooltips();",
  "applyDiscreteSliders();",
  "syncFixedFilterCells();",
  "}, 75);",
  "});",
  "window.setTimeout(function() {",
  "applyHeaderTooltips();",
  "applyDiscreteSliders();",
  "syncFixedFilterCells();",
  "}, 0);",
  "}"
))

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

.benchmark-table-wide .dataTables_scrollHead.benchmark-filter-open {
  overflow: visible !important;
  position: relative;
  z-index: 10;
}

.benchmark-table-wide table.dataTable tr > :nth-last-child(2).dtfc-fixed-right {
  box-shadow: -4px 0 4px -4px var(--bs-secondary-color);
}

.benchmark-column-guide {
  border-bottom: 1px solid var(--bs-border-color);
  border-top: 1px solid var(--bs-border-color);
  margin-top: 1rem;
  padding: 0.75rem 0;
}

.benchmark-column-guide summary {
  cursor: pointer;
  font-weight: 600;
}

.benchmark-column-guide table {
  margin-top: 0.75rem;
}

.benchmark-discrete-values {
  clear: both;
  color: var(--bs-secondary-color);
  font-size: 0.75rem;
  line-height: 1.35;
  padding-top: 0.5rem;
  white-space: normal;
}

@media (min-width: 768px) {
  .approach-guidance-table table {
    table-layout: fixed;
    width: 100%;
  }

  .approach-guidance-table col:nth-child(1) {
    width: 14%;
  }

  .approach-guidance-table col:nth-child(2) {
    width: 30%;
  }

  .approach-guidance-table col:nth-child(3) {
    width: 56%;
  }
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
runtime and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 128 GiB RAM.
The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
We recommend benchmarking a representative subset of the intended workload before committing
substantial compute resources.
</div>

Runtime is measured directly around `shapr::explain()`.
This timing excludes preparing the benchmark data and fitting the models.
Peak RAM is the highest memory used during the full benchmark run, including
loaded data, the fitted prediction model, and workers.
To limit compute time, some of the most expensive VAEAC, ARF, and timeseries
configurations were run twice; all other configurations were run three times.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks)
at GitHub.

## Key findings and practical guidance

- **`approach`: choose for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data in most tested
  settings, although regression surrogate is faster at large `x_explain` (100+).
  Regression surrogate is fastest for the mixed-data settings, and
  the categorical approach for fully categorical data. These are cost
  comparisons, not rankings of explanation quality.

- **`x_explain`: approaches differ in setup cost and cost per explained
  observation.** The approaches fall into three broad cost patterns.
  1. **Low setup, increasing cost per observation:** Gaussian, copula,
  empirical, categorical, and independence (not recommended) are attractive when explaining only
  a few observations, but their runtime increases as `x_explain` grows.
  2. **High setup, low cost per observation:** For regression surrogate,
  regression separate, and VAEAC, setup costs typically dominate while each
  additional observation adds relatively little runtime. They can therefore
  become more competitive for larger `x_explain` (although VAEAC remains
  costly).
  3. **High setup, increasing cost per observation:** Ctree, ARF, and timeseries
  combine setup costs with strong runtime growth as `x_explain` increases.

  When appropriate, explaining a representative sample can reduce runtime for
  methods whose cost increases with `x_explain` (category 1 and 3).

- **`x_train`: reducing the training set can lower runtime for most
  approaches.** Runtime changes little with `x_train` for Gaussian, regression
  separate, and independence (not recommended). It increases more clearly for empirical, copula,
  categorical, Ctree, ARF, timeseries, VAEAC, and regression surrogate.
  Subsampling `x_train` can therefore reduce runtime for these approaches (when
  statistically appropriate).

- **`r popover_note("workers")`: one worker is often fine; try four for expensive
  runs.** Parallel overhead commonly outweighs useful work for small explanation
  tasks. Heavy Gaussian, empirical, Ctree, ARF, timeseries, and
  explanation-heavy VAEAC workloads can benefit. Independence, copula,
  categorical, regression separate, and regression surrogate do not benefit at
  the tested sizes. Additional workers increase memory use, and runtime gains
  often diminish beyond four workers. Parallel VAEAC requires
  `future::multicore`.

- **`r popover_note("batching", c("min_n_batches", "max_batch_size", "max_batch_cube_size"))`:
  use batching to manage memory.** Increase the number of batches when memory is
  limiting; reduce it when repeated batch overhead dominates runtime. Fine
  batching is comparatively inexpensive for Gaussian, Ctree, and ARF, but can
  be costly for empirical, timeseries, and VAEAC. For Gaussian, copula, and
  empirical, `max_batch_cube_size` provides an additional cap on dense
  per-batch arrays.

- **`iterative`: use when no reasonable coalition budget is known.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when not even a rough fixed budget is known; otherwise, prefer a fixed run.

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: control the main
  generic costs.** Feature count, training size, and factor complexity can also
  be important, but their effects vary more by approach (see below).

- **`model` and `n_MC_samples`: prediction cost can amplify sampling cost.**
  Model prediction is repeated across coalitions and conditional samples. For
  slow prediction models, reducing `n_MC_samples` may therefore shorten runtime
  considerably; for fast models, other work may dominate.

- **Approach-specific arguments can dominate generic settings.** Examples include
  empirical AICc bandwidth selection, regression-separate learner tuning, VAEAC
  training, fine timeseries or VAEAC batches, and parallel ARF.

## Additional guidance by approach

The controls described above, including `max_n_coalitions`, `n_MC_samples`,
`x_explain`, `x_train`, workers, and batching, apply across approaches. This
section highlights only additional approach-specific controls and important
constraints that affect how those controls should be used. Reducing statistical
budgets or model complexity can change the explanation, so choose them based on
the required quality rather than cost alone. Pass approach-specific arguments
to `explain()` through `...`; see the [manual](../reference/explain.html#arg--).

::: {.approach-guidance-table}

| Approach | Additional cost control or constraint | How to reduce computational cost |
|:--|:--|:--|
| Gaussian | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Empirical | `empirical.type` | **Time:** Keep `"fixed_sigma"` unless adaptive bandwidth selection is needed; `"AICc_full"` and `"AICc_each_k"` are orders of magnitude slower. |
| Ctree | Strong sensitivity to feature and factor complexity | **Both:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible. Many factor levels are especially costly when many factor columns are present. |
| ARF | `arf.num_trees`, `arf.max_iters`; high per-worker memory | **Both:** Lower `arf.num_trees`, `arf.max_iters`, `max_n_coalitions`, or `n_MC_samples`, then verify the fitted conditional distribution.<br>**RAM:** Add `r popover_note("workers")` only after checking available memory. |
| Timeseries | High runtime overhead from fine `r popover_note("batching", c("min_n_batches", "max_batch_size"))`; strong sensitivity to feature and data structure | **Time:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible, and prefer larger batches.<br>**RAM:** Use smaller batches when needed; memory rises sharply with workers. |
| VAEAC | `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, `vaeac.latent_dim`; training-dominated runtime | **Both:** Lower `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, or `vaeac.latent_dim`, or reduce `x_train`, then verify explanation quality.<br>**Time:** Prefer larger batches; additional `r popover_note("workers")` mainly help when post-training coalition estimation, like conditional sampling and prediction, dominates runtime. |
| Independence | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Copula | Sensitivity to `x_train` | **Time:** Reduce `x_train` as well as `max_n_coalitions`, `n_MC_samples`, or `x_explain`. |
| Categorical | Limited sensitivity to `n_MC_samples` | **Time:** Prioritize reducing `max_n_coalitions`, `x_train`, or `x_explain` before reducing Monte Carlo samples. |
| Regression separate | `regression.model`, `regression.tune_values`, `regression.vfold_cv_para`; fitting-dominated runtime | **Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate; among the general controls, prioritize `max_n_coalitions`. |
| Regression surrogate | `regression.surrogate_n_comb` and regression fitting controls | **Both:** Reduce `regression.surrogate_n_comb`, `x_train`, or `max_n_coalitions`.<br>**Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate. Verify that the resulting surrogate remains adequate. |

:::

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. The retained experiments
cover `r paste(sort(unique(benchmark_results$n_features)), collapse = ", ")`
features, with available values varying by approach. Use the controls above each
column to filter: categorical controls accept one or more values, and sliders
for experiment inputs snap to values present in the results. Click headings to
sort, and use **Column visibility** to focus on relevant fields. Runtime and peak
RAM remain fixed on the right while scrolling horizontally; runtime IQR is
available under **Column visibility**. **CSV** downloads the currently displayed
table.

The default view shows the main workload inputs, approach-specific settings,
replicate count, runtime, and peak RAM.
Less frequently needed experiment details remain available under **Column visibility**.
Hover over any column name for its description.

<details class="benchmark-column-guide">
<summary>Column descriptions</summary>

Direct `explain()` arguments link to the function manual. Batching controls
link to [`get_extra_comp_args_default()`](../reference/get_extra_comp_args_default.html)
and are passed through `extra_computation_args`.

```{r benchmark-column-guide, results = "asis"}
knitr::kable(
  column_guide,
  format = "html",
  escape = FALSE,
  align = c("l", "l"),
  table.attr = 'class="table table-sm"'
)
```

</details>

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  colnames = display_name_overrides,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      list(extend = "colvis", text = "Column visibility")
    ),
    columnDefs = list(
      list(targets = hidden_column_indices, visible = FALSE)
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 0, rightColumns = 2),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE,
    initComplete = benchmark_table_callback
  )
) |>
  DT::formatRound(
    columns = c(
      "Runtime (s)",
      "Runtime IQR (s)"
    ),
    digits = 1
  ) |>
  DT::formatRound(
    columns = "Peak RAM (MiB)",
    digits = 0
  ) |>
  DT::formatRound(
    columns = "Iterations",
    digits = 1
  )
```

</div>
