---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)
benchmark_results[, approach := as.character(approach)]

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
elapsed time and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 126
GiB RAM. The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
Benchmark a representative subset of the intended workload before committing
substantial compute resources.
</div>

The headline elapsed-time measure is whole-process wall time, which includes R
startup, loading the pre-built model and data, and `explain()`. The table also
contains `explain()` wall time separately. Peak RAM covers the measured process
tree or cgroup, including workers. Most configurations have three replicates;
the expensive VAEAC and realistic ARF and timeseries blocks have two.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks).

## Key findings and practical guidance

- **Choose the approach for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data, regression
  surrogate for the mixed-data settings, and the categorical approach
  for fully categorical data. These are cost comparisons, not rankings of
  explanation quality.

- **Workload size can change the preferred approach.** Low-setup approaches
  such as Gaussian, copula, and empirical are attractive for few explanations.
  Regression surrogate, regression separate, and VAEAC have reusable setup
  costs and add little runtime as `n_explain` increases; this makes the
  regression approaches more competitive, although VAEAC remains expensive.
  (CTree and ARF also perform setup work, but their runtime grows strongly with
  `n_explain`). Large training sets favor methods with weak training-size
  sensitivity.

- **Start with one worker; try four for expensive runs.** Parallel overhead
  commonly outweighs useful work for small explanations. Heavy workloads can
  benefit, but additional workers increase memory use and gains often diminish
  beyond four.

- **The batching defaults are usually a good starting point.** Increase the
  number of batches when memory is limiting. Fine batching is comparatively
  inexpensive for Gaussian, CTree, and ARF, but costly for timeseries and
  VAEAC.

- **Use iteration for adaptive stopping, not as a shortcut.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when a suitable fixed budget is not already known.

- **Coalitions and explained observations are the clearest general cost
  drivers.** Feature count, training size, factor complexity, and Monte Carlo
  samples can also be important, but their effects vary more by approach.

- **Expensive prediction models amplify explanation cost.** Their prediction
  work is repeated across coalitions and conditional samples, so budget
  `n_MC_samples` carefully. This interaction is expected but was not measured
  directly in this benchmark.

- **Some method-specific settings dominate everything else.** Empirical AICc
  bandwidth selection, regression-separate learner tuning, VAEAC training, fine
  timeseries or VAEAC batches, and parallel ARF require explicit planning.

## Findings and guidance by approach

### Gaussian

Gaussian is inexpensive for the shared reference workload. Coalition count,
explained rows, and Monte Carlo samples increase cost, while training size has
little effect in the tested range. Fine batching reduces memory with little
overhead. Four workers help on heavy workloads, with diminishing gains at
eight and sixteen. The default dense-array cap of one million elements gave the
lowest memory without a speed penalty in the calibration and remains a sensible
default when the memory budget is unknown.

### Empirical

Empirical cost grows with training rows, coalitions, and explained rows. Fine
batching can reduce RAM, but very small batches add repeated-search overhead.
Four or eight workers help on heavy workloads before performance saturates.
Repeated AICc bandwidth selection is orders of magnitude more
expensive than a fixed bandwidth in the tested block and should be budgeted as
a first-order modeling choice.

### CTree

CTree is sensitive to feature count, coalitions, explained rows, training rows,
and mixed or factor complexity. Batching reduces peak memory with little cost in
the ordinary tested range. Heavy CTree workloads show strong parallel scaling,
but RAM increases substantially with worker-local state. Four workers
are a reasonable first parallel trial when the sequential runtime is material.

### ARF

ARF time and memory grow strongly with the main problem dimensions and complex
mixed data. More batches control memory at a modest time cost. Heavy workloads
parallelize well, but duplicated model and sampling state make the RAM cost
unusually large. Start with four workers only after confirming that
the sequential workload is expensive and that sufficient memory is available.

### Timeseries

Timeseries is sensitive to features, coalitions, explained rows, training rows,
and data structure. Its per-batch overhead is high, so prefer coarse batching
unless memory requires a finer split. Material workloads can benefit from
parallel execution, but absolute RAM use can become very large. In the retained
heavy case, eight and sixteen workers were effectively tied, leaving no reason
to use sixteen.

### VAEAC

VAEAC cost is commonly dominated by neural-network training. Training rows and
epochs are primary time drivers, while architecture, latent size, factor
complexity, and explanation dimensions can also affect time or memory. Fine
batching adds expensive repeated overhead. Training-dominated work shows
little parallel benefit; explanation-heavy work improves at four workers and
only modestly thereafter. Reusing or amortizing a trained model should be
considered before adding workers.

### Independence

Independence is inexpensive and largely insensitive to training size. Coalition
count, explained rows, and Monte Carlo samples are its relevant generic cost
drivers. Parallel overhead exceeds useful work at the retained sizes and
multiplies memory, so one worker is preferred.

### Copula

Copula is also inexpensive at the shared reference workload, but is more
sensitive to training size than independence. Coalition count and explanation
size remain important. The benchmark indicates that the retained workload does
not justify multiple workers; start with one unless a materially heavier
representative case demonstrates a benefit.

### Categorical

Categorical cost increases with training and explanation size, while Monte
Carlo count is a weak driver in the tested block. Parallel overhead outweighs
useful work at the retained size, making one worker the appropriate default.

### Regression separate

Regression separate is comparatively insensitive to generic Monte Carlo,
explanation, batching, and data.table-thread changes. Coalition count matters,
but learner and tuning choices dominate: smooth-model or XGBoost
cross-validation is much more expensive than untuned variants. The retained
workload does not benefit from shapr worker parallelism, so resource planning
should focus first on the fitting strategy.

### Regression surrogate

Regression surrogate grows with training size, coalition count, and the number
of surrogate combinations; explained rows and Monte Carlo count are weaker
cost drivers in the tested range. Its parallel experiment is unfavorable:
additional workers increase both elapsed time and RAM. Use one shapr worker
unless a materially different workload demonstrates otherwise.

## Detailed cross-approach findings

### Approach choice depends on the data and workload

For the shared numeric reference workload (12 features, 1,000 training rows,
25 explained rows, 128 maximum coalitions, 250 Monte Carlo samples, and one
worker), Gaussian is the fastest dependence-aware approach: the median
whole-process time is 4.12 seconds, compared with 4.97 seconds for copula and
7.22 seconds for empirical. Independence is similarly inexpensive, but it
deliberately ignores feature dependence and should only be used when that
approximation is appropriate. Gaussian is therefore the strongest
computational starting point for numeric data when its distributional
assumptions are suitable, but cost alone should not determine the statistical
approach.

For the four mixed-data configurations, the benchmark indicates that the
default regression-surrogate setup is fastest among the retained
dependence-aware approaches, followed by regression separate. For fully
categorical data, the dedicated categorical approach is fastest at the shared
reference workload. CTree, ARF, and VAEAC are more expensive, sometimes
substantially so, but offer different models of
the conditional distribution. These timings rank computational cost under the
tested defaults; they do not rank explanation quality.

### Explained and training observations change the ranking

For one or a few explained observations, approaches with little setup work are
favored. Gaussian, copula, and empirical are comparatively inexpensive, while
the setup cost of the regression approaches has little opportunity to
amortize. At 100 explained observations in the shared numeric sweep, regression
surrogate is slightly faster than Gaussian (6.12 versus 6.59 seconds for
the whole process) because its fitted surrogate is reused. Regression
separate and VAEAC also change little with `n_explain` in the tested sweep,
although VAEAC remains much slower in absolute terms because training
dominates. CTree, ARF, and especially timeseries grow much more strongly with
the number of explained observations.

Training size has little effect on Gaussian and independence and only a modest
effect on regression separate. It matters considerably for empirical, copula,
categorical, CTree, ARF, timeseries, VAEAC, and regression surrogate. For large
training sets, do not extrapolate from a small pilot unless the benchmark shows
weak training-size sensitivity; fit or search costs can dominate the ordinary
explanation dimensions.

### Features, coalitions, and Monte Carlo samples

Coalition count and the number of explained observations are the most
consistent generic time drivers. Feature count can additionally change the
available coalition space and approach-specific internal representations, and
is especially important for CTree, ARF, timeseries, and grouped Gaussian
runs. Factor count and the number of factor levels materially affect CTree,
ARF, and VAEAC.

`max_n_coalitions` and `n_MC_samples` should be treated as accuracy controls
with a measurable computational price. Their cost differs by approach:
Monte Carlo count is visible for Gaussian, independence, copula, and ARF, but
is weak where fitting, training, or per-batch setup dominates. This study does
not determine which values are accurate enough.

### Parallel workers

One worker is generally best for small workloads. At the ordinary tested
sizes, extra workers do not help Gaussian, independence, copula, categorical,
regression separate, or regression surrogate consistently, and always raise
RAM through worker-local state. Material Gaussian, empirical, CTree, ARF,
timeseries, and explanation-heavy VAEAC workloads can benefit from
parallelization.

When sequential runtime is material, four workers are a sensible first trial.
Measure both elapsed time and peak RAM on a representative case before trying
eight or sixteen workers: time gains commonly diminish while memory continues
to grow. Regression surrogate is the clearest warning against universal
parallel advice, since additional workers make the retained workload slower as
well as more memory-intensive.

### Batching and memory

The defaults are a reasonable starting point: `min_n_batches = 10`, together
with the one-million-element dense-array cap used by Gaussian, copula, and
empirical, limits live intermediate state without requiring users to estimate a
batch size. The Gaussian cap calibration found no speed penalty at the default
and substantially lower RAM than with larger caps.

Change batching for a measured reason. More batches can sharply reduce peak
RAM for Gaussian, CTree, ARF, timeseries, and VAEAC, but every batch adds setup
and scheduling work. Fine batching has little penalty for Gaussian, CTree, and
ARF, is expensive for empirical at the finest splits, and is particularly
costly for timeseries and VAEAC. In parallel runs, use at least as many batches
as workers; in sequential timeseries or VAEAC runs, fewer batches may be much
faster if memory permits.

### Iterative versus fixed coalitions

The iterative approach is an adaptive stopping mechanism, not a computational
shortcut when the required fixed budget is already known. In matched pairs
with the same realized coalition budget, one-iteration runs cost about the same
as fixed runs. When convergence requires several iterations, the benchmark
indicates that iterative runs are about 1.1 to 5.0 times slower because work is
performed and checked in
stages. Iteration can still save substantial work when it stops below an
otherwise conservative fixed budget; use it when the required budget is
unknown, and use a fixed budget when a suitable value has already been
established.

### Prediction-model cost

The prediction model can be a first-order cost driver. In the Gaussian model
experiment, a 500-tree depth-six XGBoost model makes explanations much slower
than a linear or standard XGBoost model and retains useful parallel scaling
through sixteen workers, whereas the cheaper models are close to practical
saturation at four.

For sampling-based approaches, increasing `n_MC_samples` also increases the
amount of conditional sampling and prediction work, so its absolute cost should
be budgeted especially carefully when model prediction is expensive. The study
varied Monte Carlo count and prediction-model complexity in separate blocks,
however, and did not directly estimate their interaction.

### Important approach-specific cost traps

Several approach settings outweigh the generic workload dimensions. Repeated
AICc bandwidth selection makes empirical orders of magnitude slower than fixed
bandwidth in the tested block. Learner choice and cross-validation dominate
regression-separate cost. Neural-network training and epoch count commonly
dominate VAEAC cost. Fine batches are unusually expensive for timeseries and
VAEAC, while parallel ARF can require very large amounts of RAM. These
choices belong in the initial resource estimate rather than being treated as
minor tuning details.

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. Use the controls above each
column to filter, click headings to sort, and use **Column visibility** to focus
on relevant fields. The first three columns remain fixed while scrolling
horizontally. **CSV** downloads the currently displayed table.

Key timing and memory columns are:

- `bash_wall_median`: median whole-process wall time in seconds;
- `wall_median`: median `explain()` wall time in seconds;
- `wall_iqr`: interquartile range of `explain()` time;
- `ram_mb_median`: median peak RAM in MB;
- `ram_mb_max`: maximum peak RAM across replicates.

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      "colvis"
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 3),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE
  )
) |>
  DT::formatRound(
    columns = c(
      "wall_median",
      "wall_iqr",
      "cpu_user_median",
      "ram_mb_median",
      "ram_mb_max",
      "bash_wall_median",
      "data_load_median",
      "n_iterations_median"
    ),
    digits = 3
  )
```

</div>
