---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

popover_notes <- list(
  workers = list(
    number = 1,
    content = paste0(
      "<code>workers</code> controls how many parallel R processes <code>shapr</code> may use. Configure it through ",
      "<code>future::plan()</code> ",
      "before calling <code>explain()</code>, for example <code>future::plan(\"multisession\", workers = 4)</code>. ",
      "The default is one worker. In a fresh R session, this is equivalent to ",
      "<code>future::plan(\"sequential\")</code> or not calling <code>future::plan()</code>."
    )
  ),
  batching = list(
    number = 2,
    content = paste0(
      "<code>min_n_batches</code> and <code>max_batch_size</code> controls the number of batches in each iteration, while ",
      "<code>max_batch_cube_size</code> limits dense per-batch arrays for Gaussian, copula, and empirical. ",
      "Pass the controls through <code>extra_computation_args</code>, ",
      "for example <code>list(min_n_batches = 20)</code>. The defaults are 10, 10, and one million, respectively. ",
      "Progress is reported once per batch, so fewer batches give less frequent updates."
    )
  )
)

popover_note <- function(key, label = key) {
  note <- popover_notes[[key]]
  if (is.null(note)) {
    stop("Unknown popover note: ", key)
  }

  formatted_labels <- paste0("<code>", htmltools::htmlEscape(label), "</code>")
  formatted_label <- if (length(formatted_labels) == 1) {
    formatted_labels
  } else if (length(formatted_labels) == 2) {
    paste(formatted_labels, collapse = " and ")
  } else {
    paste0(paste(utils::head(formatted_labels, -1), collapse = ", "), ", and ", utils::tail(formatted_labels, 1))
  }

  htmltools::HTML(sprintf(
    paste0(
      '%s<sup><a class="footnote-ref" tabindex="0" role="button" ',
      'data-bs-toggle="popover" data-bs-content="%s">[%s]</a></sup>'
    ),
    formatted_label,
    htmltools::htmlEscape(note$content, attribute = TRUE),
    note$number
  ))
}

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)
benchmark_results[, approach := as.character(approach)]

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
elapsed time and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 128 GiB RAM.
The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
We recommend benchmarking a representative subset of the intended workload before committing
substantial compute resources.
</div>

The headline elapsed-time measure is whole-process wall time, which includes R
startup, loading the pre-built model and data, and `explain()`. The table also
contains `explain()` wall time separately. Peak RAM covers the measured process
tree or cgroup, including workers. Most configurations have three replicates;
the expensive VAEAC and realistic ARF and timeseries blocks have two.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks).
at GitHub.

## Key findings and practical guidance

- **`approach`: choose for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data, regression
  surrogate for the mixed-data settings, and the categorical approach for fully
  categorical data. These are cost comparisons, not rankings of explanation
  quality.

- **`x_explain` and `x_train`: their sizes can change the preferred approach.**
  Low-setup approaches such as Gaussian, copula, and empirical are attractive
  when there are few observations to explain (few rows in `x_explain`).
  Regression surrogate, regression separate,
  and VAEAC have reusable setup costs and add little runtime as `x_explain`
  grows; this makes the regression approaches more competitive, although VAEAC
  remains expensive. Ctree and ARF also perform setup work, but their runtime still
  grows strongly with `x_explain`; hence, using a sample of the original `x_explain` may reduce their runtime.
  Large `x_train` favors methods with weak
  training-size sensitivity (Gaussian, regression separate, and independence,
  (not recommended)).

- **`r popover_note("workers")`: one worker is often fine; try four for expensive
  runs.** Parallel overhead commonly outweighs useful work for small explanation
  tasks. Heavy Gaussian, empirical, Ctree, ARF, timeseries, and
  explanation-heavy VAEAC workloads can benefit. Independence, copula,
  categorical, regression separate, and regression surrogate do not benefit at
  the tested sizes. Additional workers increase memory use, and runtime gains
  often diminish beyond four workers. Parallel VAEAC requires
  `future::multicore`.

- **`r popover_note("batching", c("min_n_batches", "max_batch_size", "max_batch_cube_size"))`:
  use batching to manage memory.** Increase the number of batches when memory is
  limiting; reduce it when repeated batch overhead dominates runtime. Fine
  batching is comparatively inexpensive for Gaussian, Ctree, and ARF, but can
  be costly for empirical, timeseries, and VAEAC. For Gaussian, copula, and
  empirical, `max_batch_cube_size` provides an additional cap on dense
  per-batch arrays.

- **`iterative`: use when no reasonable coalition budget is known.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when not even a rough fixed budget is known; otherwise, prefer a fixed run.

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: control the main
  generic costs.** Feature count, training size, and factor complexity can also
  be important, but their effects vary more by approach (see below).

- **`model` and `n_MC_samples`: prediction cost can amplify sampling cost.**
  Model prediction is repeated across coalitions and conditional samples. For
  slow prediction models, reducing `n_MC_samples` may therefore shorten runtime
  considerably; for fast models, other work may dominate.

- **Approach-specific arguments can dominate generic settings.** Examples include
  empirical AICc bandwidth selection, regression-separate learner tuning, VAEAC
  training, fine timeseries or VAEAC batches, and parallel ARF.

## Additional guidance by approach

The controls described above, including `max_n_coalitions`, `n_MC_samples`,
`x_explain`, `x_train`, workers, and batching, apply across approaches. This
section highlights only additional approach-specific controls and important
constraints that affect how those controls should be used. Reducing statistical
budgets or model complexity can change the explanation, so choose them based on
the required quality rather than cost alone. Pass approach-specific arguments
to `explain()` through `...`; see the [manual](../reference/explain.html#arg--).

| Approach | Additional cost control or constraint | How to reduce computational cost |
|:--|:--|:--|
| Gaussian | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Empirical | `empirical.type` | **Time:** Keep `"fixed_sigma"` unless adaptive bandwidth selection is needed; `"AICc_full"` and `"AICc_each_k"` are orders of magnitude slower. |
| Ctree | Strong sensitivity to feature and factor complexity | **Both:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible. Many factor levels are especially costly when many factor columns are present. |
| ARF | `arf.num_trees`, `arf.max_iters`; high per-worker memory | **Both:** Lower `arf.num_trees`, `arf.max_iters`, `max_n_coalitions`, or `n_MC_samples`, then verify the fitted conditional distribution. **RAM:** Add `r popover_note("workers")` only after checking available memory. |
| Timeseries | High runtime overhead from fine `r popover_note("batching", c("min_n_batches", "max_batch_size"))`; strong sensitivity to feature and data structure | **Time:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible, and prefer larger batches. **RAM:** Use smaller batches when needed; memory rises sharply with workers. |
| VAEAC | `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, `vaeac.latent_dim`; training-dominated runtime | **Both:** Lower `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, or `vaeac.latent_dim`, or reduce `x_train`, then verify explanation quality. **Time:** Prefer larger batches; additional `r popover_note("workers")` mainly help when post-training coalition estimation, like conditional sampling and prediction, dominates runtime. |
| Independence | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Copula | Sensitivity to `x_train` | **Time:** Reduce `x_train` as well as `max_n_coalitions`, `n_MC_samples`, or `x_explain`. |
| Categorical | Limited sensitivity to `n_MC_samples` | **Time:** Prioritize reducing `max_n_coalitions`, `x_train`, or `x_explain` before reducing Monte Carlo samples. |
| Regression separate | `regression.model`, `regression.tune_values`, `regression.vfold_cv_para`; fitting-dominated runtime | **Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate; among the general controls, prioritize `max_n_coalitions`. |
| Regression surrogate | `regression.surrogate_n_comb` and regression fitting controls | **Both:** Reduce `regression.surrogate_n_comb`, `x_train`, or `max_n_coalitions`. **Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate. Verify that the resulting surrogate remains adequate. |

## Detailed cross-approach findings

### Approach choice depends on the data and workload

For the shared numeric reference workload (12 features, 1,000 training rows,
25 explained rows, 128 maximum coalitions, 250 Monte Carlo samples, and one
worker), Gaussian is the fastest dependence-aware approach: the median
whole-process time is 4.12 seconds, compared with 4.97 seconds for copula and
7.22 seconds for empirical. Independence is similarly inexpensive, but it
deliberately ignores feature dependence and should only be used when that
approximation is appropriate. Gaussian is therefore the strongest
computational starting point for numeric data when its distributional
assumptions are suitable, but cost alone should not determine the statistical
approach.

For the four mixed-data configurations, the benchmark indicates that the
default regression-surrogate setup is fastest among the retained
dependence-aware approaches, followed by regression separate. For fully
categorical data, the dedicated categorical approach is fastest at the shared
reference workload. Ctree, ARF, and VAEAC are more expensive, sometimes
substantially so, but offer different models of
the conditional distribution. These timings rank computational cost under the
tested defaults; they do not rank explanation quality.

### Explained and training observations change the ranking

For one or a few explained observations, approaches with little setup work are
favored. Gaussian, copula, and empirical are comparatively inexpensive, while
the setup cost of the regression approaches has little opportunity to
amortize. At 100 explained observations in the shared numeric sweep, regression
surrogate is slightly faster than Gaussian (6.12 versus 6.59 seconds for
the whole process) because its fitted surrogate is reused. Regression
separate and VAEAC also change little with `n_explain` in the tested sweep,
although VAEAC remains much slower in absolute terms because training
dominates. Ctree, ARF, and especially timeseries grow much more strongly with
the number of explained observations.

Training size has little effect on Gaussian and independence and only a modest
effect on regression separate. It matters considerably for empirical, copula,
categorical, Ctree, ARF, timeseries, VAEAC, and regression surrogate. For large
training sets, do not extrapolate from a small pilot unless the benchmark shows
weak training-size sensitivity; fit or search costs can dominate the ordinary
explanation dimensions.

### Features, coalitions, and Monte Carlo samples

Coalition count and the number of explained observations are the most
consistent generic time drivers. Feature count can additionally change the
available coalition space and approach-specific internal representations, and
is especially important for Ctree, ARF, timeseries, and grouped Gaussian
runs. Factor count and the number of factor levels materially affect Ctree,
ARF, and VAEAC.

`max_n_coalitions` and `n_MC_samples` should be treated as accuracy controls
with a measurable computational price. Their cost differs by approach:
Monte Carlo count is visible for Gaussian, independence, copula, and ARF, but
is weak where fitting, training, or per-batch setup dominates. This study does
not determine which values are accurate enough.

### Parallel workers

One worker is generally best for small workloads. At the ordinary tested
sizes, extra workers do not help Gaussian, independence, copula, categorical,
regression separate, or regression surrogate consistently, and always raise
RAM through worker-local state. Material Gaussian, empirical, Ctree, ARF,
timeseries, and explanation-heavy VAEAC workloads can benefit from
parallelization.

When sequential runtime is material, four workers are a sensible first trial.
Measure both elapsed time and peak RAM on a representative case before trying
eight or sixteen workers: time gains commonly diminish while memory continues
to grow. Regression surrogate is the clearest warning against universal
parallel advice, since additional workers make the retained workload slower as
well as more memory-intensive.

### Batching and memory

The defaults are a reasonable starting point: `r popover_note("batching", "min_n_batches = 10")`, together
with the one-million-element dense-array cap used by Gaussian, copula, and
empirical, limits live intermediate state without requiring users to estimate a
batch size. The Gaussian cap calibration found no speed penalty at the default
and substantially lower RAM than with larger caps.

Change batching for a measured reason. More batches can sharply reduce peak
RAM for Gaussian, Ctree, ARF, timeseries, and VAEAC, but every batch adds setup
and scheduling work. Fine batching has little penalty for Gaussian, Ctree, and
ARF, is expensive for empirical at the finest splits, and is particularly
costly for timeseries and VAEAC. In parallel runs, use at least as many batches
as workers; in sequential timeseries or VAEAC runs, fewer batches may be much
faster if memory permits.

### Iterative versus fixed coalitions

The iterative approach is an adaptive stopping mechanism, not a computational
shortcut when the required fixed budget is already known. In matched pairs
with the same realized coalition budget, one-iteration runs cost about the same
as fixed runs. When convergence requires several iterations, the benchmark
indicates that iterative runs are about 1.1 to 5.0 times slower because work is
performed and checked in
stages. Iteration can still save substantial work when it stops below an
otherwise conservative fixed budget; use it when the required budget is
unknown, and use a fixed budget when a suitable value has already been
established.

### Prediction-model cost

The prediction model can be a first-order cost driver. In the Gaussian model
experiment, a 500-tree depth-six XGBoost model makes explanations much slower
than a linear or standard XGBoost model and retains useful parallel scaling
through sixteen workers, whereas the cheaper models are close to practical
saturation at four.

For sampling-based approaches, increasing `n_MC_samples` also increases the
amount of conditional sampling and prediction work, so its absolute cost should
be budgeted especially carefully when model prediction is expensive. The study
varied Monte Carlo count and prediction-model complexity in separate blocks,
however, and did not directly estimate their interaction.

### Important approach-specific cost traps

Several approach settings outweigh the generic workload dimensions. Repeated
AICc bandwidth selection makes empirical orders of magnitude slower than fixed
bandwidth in the tested block. Learner choice and cross-validation dominate
regression-separate cost. Neural-network training and epoch count commonly
dominate VAEAC cost. Fine batches are unusually expensive for timeseries and
VAEAC, while parallel ARF can require very large amounts of RAM. These
choices belong in the initial resource estimate rather than being treated as
minor tuning details.

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. Use the controls above each
column to filter, click headings to sort, and use **Column visibility** to focus
on relevant fields. The first three columns remain fixed while scrolling
horizontally. **CSV** downloads the currently displayed table.

Key timing and memory columns are:

- `bash_wall_median`: median whole-process wall time in seconds;
- `wall_median`: median `explain()` wall time in seconds;
- `wall_iqr`: interquartile range of `explain()` time;
- `ram_mb_median`: median peak RAM in MB;
- `ram_mb_max`: maximum peak RAM across replicates.

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      "colvis"
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 3),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE
  )
) |>
  DT::formatRound(
    columns = c(
      "wall_median",
      "wall_iqr",
      "cpu_user_median",
      "ram_mb_median",
      "ram_mb_max",
      "bash_wall_median",
      "data_load_median",
      "n_iterations_median"
    ),
    digits = 3
  )
```

</div>
