---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

popover_notes <- list(
  workers = list(
    number = 1,
    content = paste0(
      "<code>workers</code> controls how many parallel R processes <code>shapr</code> may use. Configure it through ",
      "<code>future::plan()</code> ",
      "before calling <code>explain()</code>, for example <code>future::plan(\"multisession\", workers = 4)</code>. ",
      "The default is one worker. In a fresh R session, this is equivalent to ",
      "<code>future::plan(\"sequential\")</code> or not calling <code>future::plan()</code>."
    )
  ),
  batching = list(
    number = 2,
    content = paste0(
      "<code>min_n_batches</code> and <code>max_batch_size</code> controls the number of batches in each iteration, while ",
      "<code>max_batch_cube_size</code> limits dense per-batch arrays for Gaussian, copula, and empirical. ",
      "Pass the controls through <code>extra_computation_args</code>, ",
      "for example <code>list(min_n_batches = 20)</code>. The defaults are 10, 10, and one million, respectively. ",
      "Progress is reported once per batch, so fewer batches give less frequent updates."
    )
  )
)

popover_note <- function(key, label = key) {
  note <- popover_notes[[key]]
  if (is.null(note)) {
    stop("Unknown popover note: ", key)
  }

  formatted_labels <- paste0("<code>", htmltools::htmlEscape(label), "</code>")
  formatted_label <- if (length(formatted_labels) == 1) {
    formatted_labels
  } else if (length(formatted_labels) == 2) {
    paste(formatted_labels, collapse = " and ")
  } else {
    paste0(paste(utils::head(formatted_labels, -1), collapse = ", "), ", and ", utils::tail(formatted_labels, 1))
  }

  htmltools::HTML(sprintf(
    paste0(
      '%s<sup><a class="footnote-ref" tabindex="0" role="button" ',
      'data-bs-toggle="popover" data-bs-content="%s">[%s]</a></sup>'
    ),
    formatted_label,
    htmltools::htmlEscape(note$content, attribute = TRUE),
    note$number
  ))
}

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)

categorical_columns <- c(
  "dataset",
  "backend",
  "model_variant",
  "sweep",
  "approach_args",
  "pair_key",
  "pair_role"
)
benchmark_results[, (categorical_columns) := lapply(.SD, factor), .SDcols = categorical_columns]

batch_limit_columns <- c("max_batch_size", "max_batch_cube_size")
benchmark_results[, (batch_limit_columns) := lapply(.SD, factor), .SDcols = batch_limit_columns]

headline_columns <- c("bash_wall_median", "ram_mb_median")
data.table::setcolorder(
  benchmark_results,
  c(setdiff(names(benchmark_results), headline_columns), headline_columns)
)

hidden_columns <- c(
  "max_batch_size",
  "max_batch_cube_size",
  "backend",
  "dt_threads",
  "group_size",
  "save_explanations",
  "sweep",
  "pair_key",
  "pair_role",
  "wall_median",
  "wall_iqr",
  "cpu_user_median",
  "ram_mb_max",
  "data_load_median",
  "n_iterations_median"
)
hidden_column_indices <- match(hidden_columns, names(benchmark_results)) - 1

column_tooltips <- c(
  bash_wall_median = paste(
    "Headline time: Median elapsed seconds for the fresh R process, including startup, cached data/model loading,",
    "and explain()."
  ),
  ram_mb_median = paste(
    "Headline RAM: Median peak MB for the whole process tree; cgroup peak is preferred, with process-tree polling",
    "as fallback."
  ),
  wall_median = "Median seconds spent inside explain() only.",
  wall_iqr = "Interquartile range of explain() wall time across replicates.",
  cpu_user_median = paste(
    "Median user CPU seconds for the parent R process; it does not include multisession-worker CPU."
  ),
  ram_mb_max = "Largest whole-process peak RAM observed across replicates.",
  data_load_median = "Median seconds used to load the cached dataset and fitted prediction model.",
  n_iterations_median = paste(
    "Median number of Shapley estimation iterations; normally one unless iterative estimation is used."
  ),
  n = "Number of successful measured replicates represented by the row.",
  sweep = "Name of the benchmark block that generated the configuration.",
  approach_args = "Approach-specific settings encoded as name=value pairs.",
  pair_key = "Identity of a matched iterative/fixed-budget comparison.",
  pair_role = "Role within a matched iterative/fixed-budget comparison."
)
column_tooltip_map <- paste(
  sprintf(
    "%s: %s",
    encodeString(names(column_tooltips), quote = "\""),
    encodeString(unname(column_tooltips), quote = "\"")
  ),
  collapse = ", "
)

discrete_slider_columns <- c(
  "n_features",
  "max_n_coalitions",
  "n_MC_samples",
  "min_n_batches",
  "workers",
  "dt_threads",
  "n_train",
  "n_explain",
  "group_size",
  "n",
  "n_iterations_median"
)
discrete_slider_map <- paste(
  vapply(discrete_slider_columns, function(column) {
    values <- sort(unique(benchmark_results[[column]]))
    sprintf(
      "%d: [%s]",
      match(column, names(benchmark_results)) - 1,
      paste(values, collapse = ", ")
    )
  }, character(1)),
  collapse = ", "
)
benchmark_table_callback <- DT::JS(paste0(
  "function(settings, json) {",
  "var discreteValues = {", discrete_slider_map, "};",
  "var columnTooltips = {", column_tooltip_map, "};",
  "var table = this.api();",
  "var $container = $(table.table().container());",
  "var $header = $container.find('.dataTables_scrollHead thead').first();",
  "var applyHeaderTooltips = function() {",
  "table.columns().every(function(columnIndex) {",
  "var columnName = table.settings()[0].aoColumns[columnIndex].sTitle;",
  "var tooltip = columnTooltips[columnName];",
  "if (tooltip) $(this.header()).attr('title', tooltip);",
  "});",
  "};",
  "var applyDiscreteSliders = function() {",
  "Object.keys(discreteValues).forEach(function(columnIndex) {",
  "var values = discreteValues[columnIndex];",
  "var visibleIndex = table.column.index('toVisible', Number(columnIndex));",
  "if (visibleIndex === null) return;",
  "var range = {min: values[0], max: values[values.length - 1]};",
  "values.slice(1, -1).forEach(function(value, index) {",
  "var position = 100 * (index + 1) / (values.length - 1);",
  "range[position.toFixed(6) + '%'] = value;",
  "});",
  "var $slider = $header.find('tr').eq(1).children().eq(visibleIndex).find('.noUi-target');",
  "if (!$slider.length || $slider.data('benchmarkDiscrete')) return;",
  "var currentValues = $slider.val().map(Number);",
  "$slider.noUiSlider({",
  "start: currentValues,",
  "range: range,",
  "snap: true",
  "}, true);",
  "$slider.data('benchmarkDiscrete', true);",
  "var $popup = $slider.parent();",
  "$popup.children('.benchmark-discrete-values').remove();",
  "$('<div class=\"benchmark-discrete-values\"></div>')",
  ".text('Available values: ' + values.join(', ')).appendTo($popup);",
  "});",
  "};",
  "var syncFixedFilterCells = function() {",
  "var $rows = $header.find('tr');",
  "var $labels = $rows.eq(0).children();",
  "var $filters = $rows.eq(1).children();",
  "$filters.removeClass('dtfc-fixed-left dtfc-fixed-right').css({left: '', right: '', position: '', zIndex: ''});",
  "$labels.each(function(index) {",
  "var $label = $(this);",
  "var side = $label.hasClass('dtfc-fixed-right') ? 'right' :",
  "  ($label.hasClass('dtfc-fixed-left') ? 'left' : null);",
  "if (!side) return;",
  "var $filter = $filters.eq(index);",
  "$filter.addClass('dtfc-fixed-' + side).css({position: 'sticky', zIndex: 3});",
  "$filter.css(side, $label.css(side));",
  "});",
  "};",
  "document.addEventListener('mousedown', function(event) {",
  "var $input = $(event.target).closest('.dtfc-fixed-right input');",
  "if (!$input.closest('.benchmark-table-wide').length) return;",
  "event.preventDefault();",
  "var $popup = $input.parent().next();",
  "var shouldOpen = !$popup.is(':visible');",
  "$header.find('.dtfc-fixed-right input').each(function() {",
  "var $otherPopup = $(this).parent().next();",
  "if (!$otherPopup.is($popup)) $otherPopup.hide();",
  "});",
  "if (shouldOpen) {",
  "$input.closest('.dataTables_scrollHead').addClass('benchmark-filter-open');",
  "$popup.css({left: 'auto', right: 0}).show();",
  "} else {",
  "$popup.hide();",
  "$input.closest('.dataTables_scrollHead').removeClass('benchmark-filter-open');",
  "}",
  "}, true);",
  "$(document).on('mousedown.benchmarkFilters', function(event) {",
  "if ($(event.target).closest('.dtfc-fixed-right').length) return;",
  "$header.find('.dtfc-fixed-right input').each(function() {",
  "$(this).parent().next().hide();",
  "});",
  "$header.closest('.dataTables_scrollHead').removeClass('benchmark-filter-open');",
  "});",
  "table.on('column-sizing.dt.benchmarkFilters column-visibility.dt.benchmarkFilters', function() {",
  "window.setTimeout(function() {",
  "applyHeaderTooltips();",
  "applyDiscreteSliders();",
  "syncFixedFilterCells();",
  "}, 75);",
  "});",
  "window.setTimeout(function() {",
  "applyHeaderTooltips();",
  "applyDiscreteSliders();",
  "syncFixedFilterCells();",
  "}, 0);",
  "}"
))

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

.benchmark-table-wide .dataTables_scrollHead.benchmark-filter-open {
  overflow: visible !important;
  position: relative;
  z-index: 10;
}

.benchmark-table-wide table.dataTable tr > :nth-last-child(2).dtfc-fixed-right {
  box-shadow: -4px 0 4px -4px var(--bs-secondary-color);
}

.benchmark-discrete-values {
  clear: both;
  color: var(--bs-secondary-color);
  font-size: 0.75rem;
  line-height: 1.35;
  padding-top: 0.5rem;
  white-space: normal;
}

@media (min-width: 768px) {
  .approach-guidance-table table {
    table-layout: fixed;
    width: 100%;
  }

  .approach-guidance-table col:nth-child(1) {
    width: 14%;
  }

  .approach-guidance-table col:nth-child(2) {
    width: 30%;
  }

  .approach-guidance-table col:nth-child(3) {
    width: 56%;
  }
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
elapsed time and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 128 GiB RAM.
The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
We recommend benchmarking a representative subset of the intended workload before committing
substantial compute resources.
</div>

The headline elapsed-time measure is whole-process wall time, which includes R
startup, loading the pre-built model and data, and `explain()`. The table also
contains `explain()` wall time separately. Peak RAM covers the measured process
tree or cgroup, including workers. Most configurations have three replicates;
the expensive VAEAC and realistic ARF and timeseries blocks have two.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks).
at GitHub.

## Key findings and practical guidance

- **`approach`: choose for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data, regression
  surrogate for the mixed-data settings, and the categorical approach for fully
  categorical data. These are cost comparisons, not rankings of explanation
  quality.

- **`x_explain` and `x_train`: their sizes can change the preferred approach.**
  Low-setup approaches such as Gaussian, copula, and empirical are attractive
  when there are few observations to explain (few rows in `x_explain`).
  Regression surrogate, regression separate,
  and VAEAC have reusable setup costs and add little runtime as `x_explain`
  grows; this makes the regression approaches more competitive, although VAEAC
  remains expensive. Ctree and ARF also perform setup work, but their runtime still
  grows strongly with `x_explain`; hence, using a sample of the original `x_explain` may reduce their runtime.
  Large `x_train` favors methods with weak
  training-size sensitivity (Gaussian, regression separate, and independence,
  (not recommended)).

- **`r popover_note("workers")`: one worker is often fine; try four for expensive
  runs.** Parallel overhead commonly outweighs useful work for small explanation
  tasks. Heavy Gaussian, empirical, Ctree, ARF, timeseries, and
  explanation-heavy VAEAC workloads can benefit. Independence, copula,
  categorical, regression separate, and regression surrogate do not benefit at
  the tested sizes. Additional workers increase memory use, and runtime gains
  often diminish beyond four workers. Parallel VAEAC requires
  `future::multicore`.

- **`r popover_note("batching", c("min_n_batches", "max_batch_size", "max_batch_cube_size"))`:
  use batching to manage memory.** Increase the number of batches when memory is
  limiting; reduce it when repeated batch overhead dominates runtime. Fine
  batching is comparatively inexpensive for Gaussian, Ctree, and ARF, but can
  be costly for empirical, timeseries, and VAEAC. For Gaussian, copula, and
  empirical, `max_batch_cube_size` provides an additional cap on dense
  per-batch arrays.

- **`iterative`: use when no reasonable coalition budget is known.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when not even a rough fixed budget is known; otherwise, prefer a fixed run.

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: control the main
  generic costs.** Feature count, training size, and factor complexity can also
  be important, but their effects vary more by approach (see below).

- **`model` and `n_MC_samples`: prediction cost can amplify sampling cost.**
  Model prediction is repeated across coalitions and conditional samples. For
  slow prediction models, reducing `n_MC_samples` may therefore shorten runtime
  considerably; for fast models, other work may dominate.

- **Approach-specific arguments can dominate generic settings.** Examples include
  empirical AICc bandwidth selection, regression-separate learner tuning, VAEAC
  training, fine timeseries or VAEAC batches, and parallel ARF.

## Additional guidance by approach

The controls described above, including `max_n_coalitions`, `n_MC_samples`,
`x_explain`, `x_train`, workers, and batching, apply across approaches. This
section highlights only additional approach-specific controls and important
constraints that affect how those controls should be used. Reducing statistical
budgets or model complexity can change the explanation, so choose them based on
the required quality rather than cost alone. Pass approach-specific arguments
to `explain()` through `...`; see the [manual](../reference/explain.html#arg--).

::: {.approach-guidance-table}

| Approach | Additional cost control or constraint | How to reduce computational cost |
|:--|:--|:--|
| Gaussian | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Empirical | `empirical.type` | **Time:** Keep `"fixed_sigma"` unless adaptive bandwidth selection is needed; `"AICc_full"` and `"AICc_each_k"` are orders of magnitude slower. |
| Ctree | Strong sensitivity to feature and factor complexity | **Both:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible. Many factor levels are especially costly when many factor columns are present. |
| ARF | `arf.num_trees`, `arf.max_iters`; high per-worker memory | **Both:** Lower `arf.num_trees`, `arf.max_iters`, `max_n_coalitions`, or `n_MC_samples`, then verify the fitted conditional distribution.<br>**RAM:** Add `r popover_note("workers")` only after checking available memory. |
| Timeseries | High runtime overhead from fine `r popover_note("batching", c("min_n_batches", "max_batch_size"))`; strong sensitivity to feature and data structure | **Time:** Reduce `x_train`, `max_n_coalitions`, `n_MC_samples`, or `x_explain` when possible, and prefer larger batches.<br>**RAM:** Use smaller batches when needed; memory rises sharply with workers. |
| VAEAC | `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, `vaeac.latent_dim`; training-dominated runtime | **Both:** Lower `vaeac.epochs`, `vaeac.depth`, `vaeac.width`, or `vaeac.latent_dim`, or reduce `x_train`, then verify explanation quality.<br>**Time:** Prefer larger batches; additional `r popover_note("workers")` mainly help when post-training coalition estimation, like conditional sampling and prediction, dominates runtime. |
| Independence | Weak sensitivity to `x_train` | **Time:** Prioritize reducing `max_n_coalitions`, `n_MC_samples`, or `x_explain`; reducing `x_train` helps little. |
| Copula | Sensitivity to `x_train` | **Time:** Reduce `x_train` as well as `max_n_coalitions`, `n_MC_samples`, or `x_explain`. |
| Categorical | Limited sensitivity to `n_MC_samples` | **Time:** Prioritize reducing `max_n_coalitions`, `x_train`, or `x_explain` before reducing Monte Carlo samples. |
| Regression separate | `regression.model`, `regression.tune_values`, `regression.vfold_cv_para`; fitting-dominated runtime | **Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate; among the general controls, prioritize `max_n_coalitions`. |
| Regression surrogate | `regression.surrogate_n_comb` and regression fitting controls | **Both:** Reduce `regression.surrogate_n_comb`, `x_train`, or `max_n_coalitions`.<br>**Time:** Use a simpler learner and avoid tuning or cross-validation when appropriate. Verify that the resulting surrogate remains adequate. |

:::

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. The retained experiments
cover `r paste(sort(unique(benchmark_results$n_features)), collapse = ", ")`
features, with available values varying by approach. Use the controls above each
column to filter: categorical controls accept one or more values, and sliders
for experiment inputs snap to values present in the results. Click headings to
sort, and use **Column visibility** to focus on relevant fields. The headline
time and RAM metrics remain fixed on the right while scrolling horizontally.
**CSV** downloads the currently displayed table.

The default view keeps the main workload inputs, approach-specific settings,
replicate count, and two headline outcomes. Less frequently needed execution
details and secondary outcomes remain available under **Column visibility**.
Hover over a result or metadata column name for its description.

### Result and metadata columns

| Column | Meaning |
|:--|:--|
| `bash_wall_median` | **Headline time:** median elapsed seconds for the fresh R process, including startup, cached data/model loading, and `explain()`. |
| `ram_mb_median` | **Headline RAM:** median peak MB for the whole process tree; cgroup peak is preferred, with process-tree polling as fallback. |
| `wall_median` | Median seconds spent inside `explain()` only. |
| `wall_iqr` | Interquartile range of `explain()` wall time across replicates. |
| `cpu_user_median` | Median user CPU seconds for the parent R process; it does not include multisession-worker CPU. |
| `ram_mb_max` | Largest whole-process peak RAM observed across replicates. |
| `data_load_median` | Median seconds used to load the cached dataset and fitted prediction model. |
| `n_iterations_median` | Median number of Shapley estimation iterations; normally one unless iterative estimation is used. |
| `n` | Number of successful measured replicates represented by the row. |
| `sweep` | Name of the benchmark block that generated the configuration. |
| `approach_args` | Approach-specific settings encoded as `name=value` pairs. |
| `pair_key`, `pair_role` | Identity and role of matched iterative/fixed-budget comparisons. |

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      list(extend = "colvis", text = "Column visibility")
    ),
    columnDefs = list(
      list(targets = hidden_column_indices, visible = FALSE)
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 0, rightColumns = 2),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE,
    initComplete = benchmark_table_callback
  )
) |>
  DT::formatRound(
    columns = c(
      "wall_median",
      "wall_iqr",
      "cpu_user_median",
      "ram_mb_median",
      "ram_mb_max",
      "bash_wall_median",
      "data_load_median",
      "n_iterations_median"
    ),
    digits = 3
  )
```

</div>
