---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)
benchmark_results[, approach := as.character(approach)]

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
elapsed time and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 126
GiB RAM. The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
Benchmark a representative subset of the intended workload before committing
substantial compute resources.
</div>

The headline elapsed-time measure is whole-process wall time, which includes R
startup, loading the pre-built model and data, and `explain()`. The table also
contains `explain()` wall time separately. Peak RAM covers the measured process
tree or cgroup, including workers. Most configurations have three replicates;
the expensive VAEAC and realistic ARF and timeseries blocks have two.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks).

## Key findings and practical guidance

- **`approach`: choose for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data, regression
  surrogate for the mixed-data settings, and the categorical approach for fully
  categorical data. These are cost comparisons, not rankings of explanation
  quality.

- **`x_explain` and `x_train`: their sizes can change the preferred approach.**
  Low-setup approaches such as Gaussian, copula, and empirical are attractive
  when there are few observations to explain (few rows in `x_explain`).
  Regression surrogate, regression separate,
  and VAEAC have reusable setup costs and add little runtime as `x_explain`
  grows; this makes the regression approaches more competitive, although VAEAC
  remains expensive. CTree and ARF also perform setup work, but their runtime still
  grows strongly with `x_explain`; hence, using a sample of the original `x_explain` may reduce their runtime.
  Large `x_train` favors methods with weak
  training-size sensitivity (Gaussian, regression separate, and independence,
  (not recommended)).

- ** `workers`: one is often fine; try e.g.~four for expensive runs.** Parallel overhead
  commonly outweighs useful work for small explanation tasks. Heavy tasks can
  benefit, but additional workers increase memory use and gains often diminish
  beyond four.

- **`min_n_batches`: use batching to manage memory.** Increase the number of
  batches when memory is limiting; reduce it when repeated batch overhead
  dominates runtime. Fine batching is comparatively inexpensive for Gaussian,
  CTree, and ARF, but costly for timeseries and VAEAC.

- **`iterative`: use adaptive stopping, not a shortcut.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when not even a rough fixed budget is known.

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: control the main
  generic costs.** Feature count, training size, and factor complexity can also
  be important, but their effects vary more by approach.

- **`model` and `n_MC_samples`: budget expensive predictions carefully.** Model
  prediction is repeated across coalitions and conditional samples. This
  interaction is expected but was not measured directly in this benchmark.

- **Approach-specific arguments can dominate generic settings.** Empirical AICc
  bandwidth selection, regression-separate learner tuning, VAEAC training, fine
  timeseries or VAEAC batches, and parallel ARF require explicit planning.

`workers` is configured through `future::plan()`. The default is `workers = 1`,
which means that `shapr` does not use parallel workers. For example, use
`future::plan("multisession", workers = 4)` to try four workers.
`min_n_batches` is an element of `extra_computation_args` passed to `explain()`;
it defaults to 10 and sets the minimum number of batches per iteration. For
example, use `extra_computation_args = list(min_n_batches = 20)`. With
`progressr`, `shapr` reports progress once per batch, so fewer batches also give
fewer progress updates.

## Findings and guidance by approach

### Gaussian

overhead. Four workers help on heavy workloads, with diminishing gains at
- **`max_n_coalitions` and `n_MC_samples`: spend the accuracy budget carefully.**
  Both increase runtime; training size has little effect in the tested range.

- **`min_n_batches` and `max_batch_cube_size`: use them to limit RAM.** Fine
  batching had little time penalty, and the default dense-array cap of one
  million elements was the best tested starting point when memory is unknown.

- **`workers`: try four only for a material workload.** Further gains diminished
  at eight and sixteen workers.

### Empirical

- **`empirical.type`: use `"fixed_sigma"` unless repeated AICc bandwidth
  selection is required.** `"AICc_full"` and `"AICc_each_k"` were orders of
  magnitude slower in the tested block; set `empirical.fixed_sigma` explicitly
  when using the fixed option.

- **`min_n_batches` and `workers`: increase batches only to meet a memory
  limit, then try four or eight workers for a heavy workload.** Very fine
  batching adds repeated search overhead, while parallel gains eventually
  saturate.

### CTree

- **`max_n_coalitions`, `n_MC_samples`, and the tree controls
  (`ctree.mincriterion`, `ctree.minsplit`, and `ctree.minbucket`): plan these
  together.** CTree cost is sensitive to feature count, coalition count,
  training size, explained rows, and mixed or factor complexity.

- **`min_n_batches` and `workers`: use more batches to reduce RAM, then try
  four workers for a heavy workload.** Batching had little overhead in the
  tested range, but worker-local state increases memory substantially.

### ARF

- **`arf.num_trees`, `arf.max_iters`, `max_n_coalitions`, and `n_MC_samples`:
  budget the main ARF workload deliberately.** Time and memory grow strongly
  with these settings and with complex mixed data.

- **`min_n_batches`, `workers`, and `arf.parallel_gen`: treat parallelism as a
  memory trade-off.** More batches reduce RAM at a modest time cost. Heavy
  workloads can scale with workers, but avoid nested parallelism and confirm
  that enough RAM is available before starting with four workers.

### Timeseries

to use sixteen.
- **`min_n_batches`: keep batches coarse unless memory requires a finer
  split.** Timeseries has high per-batch overhead, especially as feature,
  coalition, training, and explanation sizes grow.

- **`workers`: benchmark a material case and stop at the first saturation
  point.** Parallel runs can help, but absolute RAM can become very large; the
  retained heavy case showed no gain from sixteen over eight workers.

### VAEAC

- **`vaeac.epochs`: this is usually the first cost control.** Training rows,
  epochs, and architecture settings such as `vaeac.depth`, `vaeac.width`, and
  `vaeac.latent_dim` can dominate elapsed time or RAM; consider amortizing a
  trained model before adding workers.

- **`min_n_batches` and `workers`: avoid fine batching and use parallelism
  selectively.** Fine batching has high repeated overhead. Training-dominated
  work showed little parallel benefit, while explanation-heavy work improved at
  four workers. More than one worker requires `future::multicore` where
  forking is available; otherwise, run sequentially.

### Independence

drivers. Parallel overhead exceeds useful work at the retained sizes and
- **`max_n_coalitions` and `n_MC_samples`: these are the relevant cost
  controls.** Independence is inexpensive and largely insensitive to training
  size in the tested range.

- **`workers`: keep the default of one.** Parallel overhead exceeded useful
  work at the retained sizes and multiplied memory.

### Copula

- **`max_n_coalitions`, `n_MC_samples`, and `min_n_batches`: change these only
  for an accuracy or memory requirement.** Copula is inexpensive at the shared
  reference workload, but grows more with training size than independence.

- **`workers`: start with one.** The retained workload did not justify
  parallel workers; benchmark a materially heavier representative case before
  increasing them.

### Categorical

- **`max_n_coalitions` and `min_n_batches`: use these to manage the main
  workload and memory trade-off.** Cost increased with training and explanation
  size, while `n_MC_samples` was a weak driver in the tested block.

- **`workers`: keep the default of one for ordinary workloads.** Parallel
  overhead outweighed useful work at the retained size.

### Regression separate

- **`regression.model`, `regression.tune_values`, and
  `regression.vfold_cv_para`: choose the fitting strategy before tuning generic
  settings.** Learner choice and cross-validation dominated cost, while
  `max_n_coalitions` remained the main generic driver.

- **`workers`: keep the default of one for the retained workload.** Focus
  resource planning on the fitting strategy rather than `shapr` parallelism.

### Regression surrogate

- **`regression.surrogate_n_comb`: use the smallest value that supports an
  adequate surrogate model.** More surrogate combinations, training rows, and
  `max_n_coalitions` increase cost; explained rows and `n_MC_samples` were
  weaker drivers in the tested range.

- **`regression.model`, `regression.tune_values`, and `workers`: keep fitting
  simple before attempting parallelism.** Additional workers increased both
  elapsed time and RAM in the retained experiment.

## Detailed cross-approach findings

### Approach choice depends on the data and workload

For the shared numeric reference workload (12 features, 1,000 training rows,
25 explained rows, 128 maximum coalitions, 250 Monte Carlo samples, and one
worker), Gaussian is the fastest dependence-aware approach: the median
whole-process time is 4.12 seconds, compared with 4.97 seconds for copula and
7.22 seconds for empirical. Independence is similarly inexpensive, but it
deliberately ignores feature dependence and should only be used when that
approximation is appropriate. Gaussian is therefore the strongest
computational starting point for numeric data when its distributional
assumptions are suitable, but cost alone should not determine the statistical
approach.

For the four mixed-data configurations, the benchmark indicates that the
default regression-surrogate setup is fastest among the retained
dependence-aware approaches, followed by regression separate. For fully
categorical data, the dedicated categorical approach is fastest at the shared
reference workload. CTree, ARF, and VAEAC are more expensive, sometimes
substantially so, but offer different models of
the conditional distribution. These timings rank computational cost under the
tested defaults; they do not rank explanation quality.

### Explained and training observations change the ranking

For one or a few explained observations, approaches with little setup work are
favored. Gaussian, copula, and empirical are comparatively inexpensive, while
the setup cost of the regression approaches has little opportunity to
amortize. At 100 explained observations in the shared numeric sweep, regression
surrogate is slightly faster than Gaussian (6.12 versus 6.59 seconds for
the whole process) because its fitted surrogate is reused. Regression
separate and VAEAC also change little with `n_explain` in the tested sweep,
although VAEAC remains much slower in absolute terms because training
dominates. CTree, ARF, and especially timeseries grow much more strongly with
the number of explained observations.

Training size has little effect on Gaussian and independence and only a modest
effect on regression separate. It matters considerably for empirical, copula,
categorical, CTree, ARF, timeseries, VAEAC, and regression surrogate. For large
training sets, do not extrapolate from a small pilot unless the benchmark shows
weak training-size sensitivity; fit or search costs can dominate the ordinary
explanation dimensions.

### Features, coalitions, and Monte Carlo samples

Coalition count and the number of explained observations are the most
consistent generic time drivers. Feature count can additionally change the
available coalition space and approach-specific internal representations, and
is especially important for CTree, ARF, timeseries, and grouped Gaussian
runs. Factor count and the number of factor levels materially affect CTree,
ARF, and VAEAC.

`max_n_coalitions` and `n_MC_samples` should be treated as accuracy controls
with a measurable computational price. Their cost differs by approach:
Monte Carlo count is visible for Gaussian, independence, copula, and ARF, but
is weak where fitting, training, or per-batch setup dominates. This study does
not determine which values are accurate enough.

### Parallel workers

One worker is generally best for small workloads. At the ordinary tested
sizes, extra workers do not help Gaussian, independence, copula, categorical,
regression separate, or regression surrogate consistently, and always raise
RAM through worker-local state. Material Gaussian, empirical, CTree, ARF,
timeseries, and explanation-heavy VAEAC workloads can benefit from
parallelization.

When sequential runtime is material, four workers are a sensible first trial.
Measure both elapsed time and peak RAM on a representative case before trying
eight or sixteen workers: time gains commonly diminish while memory continues
to grow. Regression surrogate is the clearest warning against universal
parallel advice, since additional workers make the retained workload slower as
well as more memory-intensive.

### Batching and memory

The defaults are a reasonable starting point: `min_n_batches = 10`, together
with the one-million-element dense-array cap used by Gaussian, copula, and
empirical, limits live intermediate state without requiring users to estimate a
batch size. The Gaussian cap calibration found no speed penalty at the default
and substantially lower RAM than with larger caps.

Change batching for a measured reason. More batches can sharply reduce peak
RAM for Gaussian, CTree, ARF, timeseries, and VAEAC, but every batch adds setup
and scheduling work. Fine batching has little penalty for Gaussian, CTree, and
ARF, is expensive for empirical at the finest splits, and is particularly
costly for timeseries and VAEAC. In parallel runs, use at least as many batches
as workers; in sequential timeseries or VAEAC runs, fewer batches may be much
faster if memory permits.

### Iterative versus fixed coalitions

The iterative approach is an adaptive stopping mechanism, not a computational
shortcut when the required fixed budget is already known. In matched pairs
with the same realized coalition budget, one-iteration runs cost about the same
as fixed runs. When convergence requires several iterations, the benchmark
indicates that iterative runs are about 1.1 to 5.0 times slower because work is
performed and checked in
stages. Iteration can still save substantial work when it stops below an
otherwise conservative fixed budget; use it when the required budget is
unknown, and use a fixed budget when a suitable value has already been
established.

### Prediction-model cost

The prediction model can be a first-order cost driver. In the Gaussian model
experiment, a 500-tree depth-six XGBoost model makes explanations much slower
than a linear or standard XGBoost model and retains useful parallel scaling
through sixteen workers, whereas the cheaper models are close to practical
saturation at four.

For sampling-based approaches, increasing `n_MC_samples` also increases the
amount of conditional sampling and prediction work, so its absolute cost should
be budgeted especially carefully when model prediction is expensive. The study
varied Monte Carlo count and prediction-model complexity in separate blocks,
however, and did not directly estimate their interaction.

### Important approach-specific cost traps

Several approach settings outweigh the generic workload dimensions. Repeated
AICc bandwidth selection makes empirical orders of magnitude slower than fixed
bandwidth in the tested block. Learner choice and cross-validation dominate
regression-separate cost. Neural-network training and epoch count commonly
dominate VAEAC cost. Fine batches are unusually expensive for timeseries and
VAEAC, while parallel ARF can require very large amounts of RAM. These
choices belong in the initial resource estimate rather than being treated as
minor tuning details.

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. Use the controls above each
column to filter, click headings to sort, and use **Column visibility** to focus
on relevant fields. The first three columns remain fixed while scrolling
horizontally. **CSV** downloads the currently displayed table.

Key timing and memory columns are:

- `bash_wall_median`: median whole-process wall time in seconds;
- `wall_median`: median `explain()` wall time in seconds;
- `wall_iqr`: interquartile range of `explain()` time;
- `ram_mb_median`: median peak RAM in MB;
- `ram_mb_max`: maximum peak RAM across replicates.

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      "colvis"
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 3),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE
  )
) |>
  DT::formatRound(
    columns = c(
      "wall_median",
      "wall_iqr",
      "cpu_user_median",
      "ram_mb_median",
      "ram_mb_max",
      "bash_wall_median",
      "data_load_median",
      "n_iterations_median"
    ),
    digits = 3
  )
```

</div>
