---
title: "Computational cost benchmarks"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    fig_caption: yes
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Computational cost benchmarks}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 80
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

popover_notes <- list(
  workers = list(
    number = 1,
    content = paste0(
      "<code>workers</code> controls how many parallel R processes <code>shapr</code> may use, typically allowing ",
      "that many CPU cores or hardware threads to work concurrently. Configure it through <code>future::plan()</code> ",
      "before calling <code>explain()</code>, for example <code>future::plan(\"multisession\", workers = 4)</code>. ",
      "The default is one worker. In a fresh R session, this is equivalent to ",
      "<code>future::plan(\"sequential\")</code> or not calling <code>future::plan()</code>."
    )
  ),
  batching = list(
    number = 2,
    content = paste0(
      "These controls split coalition estimates within each iteration into batches. <code>min_n_batches</code> sets ",
      "the minimum number of batches, <code>max_batch_size</code> limits the coalitions processed at once, and ",
      "<code>max_batch_cube_size</code> limits dense per-batch arrays for Gaussian, copula, and empirical. Smaller ",
      "batches generally reduce peak RAM but add overhead. Pass the controls through <code>extra_computation_args</code>, ",
      "for example <code>list(min_n_batches = 20)</code>. The defaults are 10, 10, and one million, respectively."
    )
  )
)

popover_note <- function(key, label = key) {
  note <- popover_notes[[key]]
  if (is.null(note)) {
    stop("Unknown popover note: ", key)
  }

  htmltools::HTML(sprintf(
    paste0(
      '<code>%s</code><sup><a class="footnote-ref" tabindex="0" role="button" ',
      'data-bs-toggle="popover" data-bs-content="%s">[%s]</a></sup>'
    ),
    htmltools::htmlEscape(label),
    htmltools::htmlEscape(note$content, attribute = TRUE),
    note$number
  ))
}

approaches <- c(
  "gaussian",
  "empirical",
  "ctree",
  "arf",
  "timeseries",
  "vaeac",
  "independence",
  "copula",
  "categorical",
  "regression_separate",
  "regression_surrogate"
)

result_roots <- c(
  file.path("benchmarks", "results"),
  file.path("..", "benchmarks", "results")
)
result_root <- result_roots[dir.exists(result_roots)][1]
if (is.na(result_root)) {
  stop("Could not locate the benchmark result directory.")
}

summary_files <- file.path(result_root, approaches, "summary.csv")
if (!all(file.exists(summary_files))) {
  stop("One or more benchmark summary files are missing.")
}

benchmark_results <- data.table::rbindlist(
  lapply(summary_files, data.table::fread),
  fill = TRUE,
  use.names = TRUE
)
benchmark_results[, approach := factor(approach, levels = approaches)]
data.table::setorder(benchmark_results, approach, sweep)
benchmark_results[, approach := as.character(approach)]

n_runs <- sum(benchmark_results$n)
n_configurations <- nrow(benchmark_results)
```

<style>
.benchmark-callout {
  border-left: 4px solid var(--bs-primary);
  background: var(--bs-secondary-bg);
  padding: 1rem 1.25rem;
  margin: 1.5rem 0;
}

.benchmark-table-wide {
  width: min(1320px, calc(100vw - 1.5rem));
  margin-left: -0.75rem;
  padding: 0 0.5rem 1rem;
  background: var(--bs-body-bg);
  position: relative;
  z-index: 2;
}

.benchmark-table-wide .dataTables_wrapper {
  width: 100%;
}

.benchmark-table-wide .dt-buttons,
.benchmark-table-wide .dataTables_filter {
  margin-bottom: 0.75rem;
}

@media (max-width: 576px) {
  .benchmark-table-wide {
    width: calc(100vw - 1rem);
  }
}
</style>

## What this study measures

This benchmark study measures the **computational cost** of `shapr::explain()`:
elapsed time and peak RAM. It covers `r format(n_runs, big.mark = ",")` successful
runs across `r format(n_configurations, big.mark = ",")` configurations and all
11 built-in estimation approaches. Each approach is studied through controlled
blocks that vary one or a few workload dimensions around a shared baseline.

The main workload dimensions are the number of features, coalitions, Monte Carlo
samples, training observations, observations to explain, batches, and parallel
workers. The study also includes relevant approach-specific choices, such as
factor complexity, empirical bandwidth selection, regression learners, and
VAEAC architecture and training settings.

<div class="benchmark-callout">
<strong>This is a cost study, not an accuracy study.</strong> Coalition and Monte
Carlo budgets affect runtime and memory, but these results do not establish which
budgets are accurate enough for a particular explanation problem.
</div>

All measurements were collected on one Linux host running Ubuntu 24.04, with an
AMD Ryzen Threadripper 1950X processor (16 physical cores, 32 threads) and 128 GiB RAM.
The study used synthetic numeric, mixed, or categorical data. Absolute
time and RAM depend on the machine, operating system, prediction model, and
data. The most transferable evidence is therefore the relative change within a
controlled benchmark block, rather than a direct prediction of runtime on
another machine.

<div class="benchmark-callout">
<strong>Treat these results as rough planning guidelines.</strong> We provide no
guarantee that the reported timings, memory use, approach ordering, or scaling
patterns will generalize to another machine, dataset, model, or package version.
We recommend benchmarking a representative subset of the intended workload before committing
substantial compute resources.
</div>

The headline elapsed-time measure is whole-process wall time, which includes R
startup, loading the pre-built model and data, and `explain()`. The table also
contains `explain()` wall time separately. Peak RAM covers the measured process
tree or cgroup, including workers. Most configurations have three replicates;
the expensive VAEAC and realistic ARF and timeseries blocks have two.

The full design, execution framework, committed CSV files, and numerical audit
are available in the
[benchmark directory](https://github.com/NorskRegnesentral/shapr/tree/master/benchmarks).
at GitHub.

## Key findings and practical guidance

- **`approach`: choose for the data, then consider cost.** Gaussian is the
  fastest dependence-aware method for purely numeric data, regression
  surrogate for the mixed-data settings, and the categorical approach for fully
  categorical data. These are cost comparisons, not rankings of explanation
  quality.

- **`x_explain` and `x_train`: their sizes can change the preferred approach.**
  Low-setup approaches such as Gaussian, copula, and empirical are attractive
  when there are few observations to explain (few rows in `x_explain`).
  Regression surrogate, regression separate,
  and VAEAC have reusable setup costs and add little runtime as `x_explain`
  grows; this makes the regression approaches more competitive, although VAEAC
  remains expensive. Ctree and ARF also perform setup work, but their runtime still
  grows strongly with `x_explain`; hence, using a sample of the original `x_explain` may reduce their runtime.
  Large `x_train` favors methods with weak
  training-size sensitivity (Gaussian, regression separate, and independence,
  (not recommended)).

- **`r popover_note("workers")`: one is often fine; try e.g. four for expensive
  runs.** This controls how many R worker processes, and therefore CPU cores or
  hardware threads, can run `shapr` computations concurrently. Parallel overhead
  commonly outweighs useful work for small explanation tasks. Heavy tasks can
  benefit, but additional workers increase memory use, and in our experiments
  the gains often diminish beyond four workers.

- **`r popover_note("batching", "min_n_batches")`: use batching to manage memory.** Increase the
  number of batches when memory is limiting; reduce it when repeated batch
  overhead dominates runtime. Fine batching is comparatively inexpensive for
  Gaussian, Ctree, and ARF, but costly for timeseries and VAEAC.

- **`iterative`: use adaptive stopping, not a shortcut.** At the same final
  coalition budget, iterative execution costs about the same as a fixed run if
  it stops after one iteration and more if it needs several. It is most useful
  when not even a rough fixed budget is known.

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: control the main
  generic costs.** Feature count, training size, and factor complexity can also
  be important, but their effects vary more by approach.

- **`model` and `n_MC_samples`: budget expensive predictions carefully.** Model
  prediction is repeated across coalitions and conditional samples. This
  interaction is expected but was not measured directly in this benchmark.

- **Approach-specific arguments can dominate generic settings.** Empirical AICc
  bandwidth selection, regression-separate learner tuning, VAEAC training, fine
  timeseries or VAEAC batches, and parallel ARF require explicit planning.

## Findings and guidance by approach

The settings below are the most useful cost controls for each approach. Reducing
the data, coalition, Monte Carlo, or model-training budgets can also affect the
statistical result, so choose those values based on the required explanation
quality rather than cost alone.

### Gaussian

- **`max_n_coalitions` and `n_MC_samples`: reduce them to shorten runs.** The
  number of rows in `x_explain` is another important runtime control, whereas
  reducing `x_train` helps little.

- **`r popover_note("batching", "min_n_batches")`, `max_batch_size`, and
  `max_batch_cube_size`: use smaller batches to lower peak RAM.** Gaussian pays little time penalty for
  finer batches. For a heavy run, setting `r popover_note("workers", "workers = 4")` can reduce
  elapsed time, but additional workers give smaller gains and continue to
  increase RAM.

### Empirical

- **`empirical.type`: keep the default `"fixed_sigma"` for the fastest
  option.** Pass this argument directly to `explain()` through `...`; see the
  [manual](../reference/explain.html#arg--). Switching to `"AICc_full"` or `"AICc_each_k"`
  makes the tested runs orders of magnitude slower. Use these modes only when
  their adaptive bandwidth selection is needed.

- **`r popover_note("batching", "min_n_batches")` and `max_batch_size`: use
  smaller batches to lower peak RAM, but avoid more batches than memory requires.** Fine batching
  repeats expensive searches. For a heavy run, try `r popover_note("workers", "workers = 4")`;
  increase to eight only when a representative run still becomes faster and
  memory allows it.

### Ctree

- **`max_n_coalitions`, `n_MC_samples`, `x_train`, and `x_explain`: reduce
  these inputs to shorten runs and lower RAM.** Ctree is particularly sensitive
  to the number of features and to mixed or factor data, which are usually
  properties of the explanation problem rather than tuning controls.

- **`r popover_note("batching", "min_n_batches")` and `max_batch_size`: use
  smaller batches to lower peak RAM.** Ctree pays little time penalty for finer batches. Setting
  `r popover_note("workers", "workers = 4")` can substantially shorten a heavy run, but each
  worker adds considerable memory use.

### ARF

- **`arf.num_trees`, `arf.max_iters`, `max_n_coalitions`, and `n_MC_samples`:
  reduce them to make ARF faster and less memory-hungry.** Pass the ARF arguments directly to
  `explain()` through `...`; see the [manual](../reference/explain.html#arg--). Lower ARF
  training settings can change the fitted conditional distribution, so verify
  that they remain adequate for the data.

- **`r popover_note("batching", "min_n_batches")` and `max_batch_size`: use
  smaller batches to lower peak RAM at a modest time cost.** Setting
  `r popover_note("workers", "workers = 4")` can
  shorten a heavy run, but ARF has an unusually large per-worker memory cost;
  increase the worker count only after measuring available RAM.

### Timeseries

- **`max_n_coalitions`, `n_MC_samples`, `x_train`, and `x_explain`: reduce
  them to shorten runs.** Feature count and data structure also strongly affect
  cost but are generally fixed by the explanation problem.

- **`r popover_note("batching", "min_n_batches")` and `max_batch_size`: prefer
  larger batches for speed and smaller batches only when you need lower peak RAM.** Per-batch
  overhead is high. A heavy run can benefit from more workers, but RAM grows
  sharply and increasing from eight to sixteen workers provides no benefit in
  the retained case.

### VAEAC

- **`vaeac.epochs`, `vaeac.depth`, `vaeac.width`, and `vaeac.latent_dim`:
  reduce them to shorten training and potentially lower RAM.** Pass these arguments directly to
  `explain()` through `...`; see the [manual](../reference/explain.html#arg--).
  These settings control model capacity or training, so verify explanation
  quality after changing them. Reducing the rows in `x_train` also speeds up
  training.

- **`r popover_note("batching", "min_n_batches")` and `max_batch_size`: prefer
  larger batches for speed and smaller batches only when you need lower peak RAM.** More workers
  do little for training-dominated runs; `r popover_note("workers", "workers = 4")` can help when
  explanation work dominates. Parallel VAEAC requires `future::multicore`, so
  run sequentially where forking is unavailable.

### Independence

- **`max_n_coalitions`, `n_MC_samples`, and `x_explain`: reduce them to shorten
  runs.** Reducing `x_train` provides little benefit in the tested range.

- **`r popover_note("workers")`: keep the default of one.** Additional workers make the
  retained workloads no faster and increase peak RAM.

### Copula

- **`max_n_coalitions`, `n_MC_samples`, `x_train`, and `x_explain`: reduce
  them to shorten runs.** Unlike Gaussian and independence, copula also becomes
  more expensive as the training data grows.

- **`r popover_note("workers")`: keep the default of one for ordinary workloads.**
  Additional workers provide no benefit at the retained size and increase peak
  RAM.

### Categorical

- **`max_n_coalitions`, `x_train`, and `x_explain`: reduce them to shorten
  runs.** Lowering `n_MC_samples` provides comparatively little benefit in the
  tested range.

- **`r popover_note("workers")`: keep the default of one for ordinary workloads.**
  Additional workers make the retained workloads no faster and increase peak
  RAM.

### Regression separate

- **`regression.model`, `regression.tune_values`, and
  `regression.vfold_cv_para`: use a simpler learner and avoid cross-validation
  to shorten runs.** Pass these arguments directly to `explain()` through `...`;
  see the [manual](../reference/explain.html#arg--). Learner fitting and tuning dominate cost,
  while reducing `max_n_coalitions` is the most effective generic adjustment.

- **`r popover_note("workers")`: keep the default of one.** Additional `shapr` workers
  provide no benefit for the retained workload; reduce model-fitting work
  instead.

### Regression surrogate

- **`regression.surrogate_n_comb`: reduce it from its default to shorten runs
  and lower RAM.** Pass this argument directly to `explain()` through `...`; see the
  [manual](../reference/explain.html#arg--). Also reduce `x_train` or
  `max_n_coalitions` when the explanation design allows it. Check that the
  resulting surrogate model remains adequate.

- **`regression.model`, `regression.tune_values`, and
  `regression.vfold_cv_para`: use a simpler learner and avoid cross-validation
  for faster fitting.** Pass these arguments directly to `explain()` through `...`;
  see the [manual](../reference/explain.html#arg--). Keep
  `r popover_note("workers", "workers = 1")`;
  additional workers make the retained workloads slower and more
  memory-hungry.

## Detailed cross-approach findings

### Approach choice depends on the data and workload

For the shared numeric reference workload (12 features, 1,000 training rows,
25 explained rows, 128 maximum coalitions, 250 Monte Carlo samples, and one
worker), Gaussian is the fastest dependence-aware approach: the median
whole-process time is 4.12 seconds, compared with 4.97 seconds for copula and
7.22 seconds for empirical. Independence is similarly inexpensive, but it
deliberately ignores feature dependence and should only be used when that
approximation is appropriate. Gaussian is therefore the strongest
computational starting point for numeric data when its distributional
assumptions are suitable, but cost alone should not determine the statistical
approach.

For the four mixed-data configurations, the benchmark indicates that the
default regression-surrogate setup is fastest among the retained
dependence-aware approaches, followed by regression separate. For fully
categorical data, the dedicated categorical approach is fastest at the shared
reference workload. Ctree, ARF, and VAEAC are more expensive, sometimes
substantially so, but offer different models of
the conditional distribution. These timings rank computational cost under the
tested defaults; they do not rank explanation quality.

### Explained and training observations change the ranking

For one or a few explained observations, approaches with little setup work are
favored. Gaussian, copula, and empirical are comparatively inexpensive, while
the setup cost of the regression approaches has little opportunity to
amortize. At 100 explained observations in the shared numeric sweep, regression
surrogate is slightly faster than Gaussian (6.12 versus 6.59 seconds for
the whole process) because its fitted surrogate is reused. Regression
separate and VAEAC also change little with `n_explain` in the tested sweep,
although VAEAC remains much slower in absolute terms because training
dominates. Ctree, ARF, and especially timeseries grow much more strongly with
the number of explained observations.

Training size has little effect on Gaussian and independence and only a modest
effect on regression separate. It matters considerably for empirical, copula,
categorical, Ctree, ARF, timeseries, VAEAC, and regression surrogate. For large
training sets, do not extrapolate from a small pilot unless the benchmark shows
weak training-size sensitivity; fit or search costs can dominate the ordinary
explanation dimensions.

### Features, coalitions, and Monte Carlo samples

Coalition count and the number of explained observations are the most
consistent generic time drivers. Feature count can additionally change the
available coalition space and approach-specific internal representations, and
is especially important for Ctree, ARF, timeseries, and grouped Gaussian
runs. Factor count and the number of factor levels materially affect Ctree,
ARF, and VAEAC.

`max_n_coalitions` and `n_MC_samples` should be treated as accuracy controls
with a measurable computational price. Their cost differs by approach:
Monte Carlo count is visible for Gaussian, independence, copula, and ARF, but
is weak where fitting, training, or per-batch setup dominates. This study does
not determine which values are accurate enough.

### Parallel workers

One worker is generally best for small workloads. At the ordinary tested
sizes, extra workers do not help Gaussian, independence, copula, categorical,
regression separate, or regression surrogate consistently, and always raise
RAM through worker-local state. Material Gaussian, empirical, Ctree, ARF,
timeseries, and explanation-heavy VAEAC workloads can benefit from
parallelization.

When sequential runtime is material, four workers are a sensible first trial.
Measure both elapsed time and peak RAM on a representative case before trying
eight or sixteen workers: time gains commonly diminish while memory continues
to grow. Regression surrogate is the clearest warning against universal
parallel advice, since additional workers make the retained workload slower as
well as more memory-intensive.

### Batching and memory

The defaults are a reasonable starting point: `r popover_note("batching", "min_n_batches = 10")`, together
with the one-million-element dense-array cap used by Gaussian, copula, and
empirical, limits live intermediate state without requiring users to estimate a
batch size. The Gaussian cap calibration found no speed penalty at the default
and substantially lower RAM than with larger caps.

Change batching for a measured reason. More batches can sharply reduce peak
RAM for Gaussian, Ctree, ARF, timeseries, and VAEAC, but every batch adds setup
and scheduling work. Fine batching has little penalty for Gaussian, Ctree, and
ARF, is expensive for empirical at the finest splits, and is particularly
costly for timeseries and VAEAC. In parallel runs, use at least as many batches
as workers; in sequential timeseries or VAEAC runs, fewer batches may be much
faster if memory permits.

### Iterative versus fixed coalitions

The iterative approach is an adaptive stopping mechanism, not a computational
shortcut when the required fixed budget is already known. In matched pairs
with the same realized coalition budget, one-iteration runs cost about the same
as fixed runs. When convergence requires several iterations, the benchmark
indicates that iterative runs are about 1.1 to 5.0 times slower because work is
performed and checked in
stages. Iteration can still save substantial work when it stops below an
otherwise conservative fixed budget; use it when the required budget is
unknown, and use a fixed budget when a suitable value has already been
established.

### Prediction-model cost

The prediction model can be a first-order cost driver. In the Gaussian model
experiment, a 500-tree depth-six XGBoost model makes explanations much slower
than a linear or standard XGBoost model and retains useful parallel scaling
through sixteen workers, whereas the cheaper models are close to practical
saturation at four.

For sampling-based approaches, increasing `n_MC_samples` also increases the
amount of conditional sampling and prediction work, so its absolute cost should
be budgeted especially carefully when model prediction is expensive. The study
varied Monte Carlo count and prediction-model complexity in separate blocks,
however, and did not directly estimate their interaction.

### Important approach-specific cost traps

Several approach settings outweigh the generic workload dimensions. Repeated
AICc bandwidth selection makes empirical orders of magnitude slower than fixed
bandwidth in the tested block. Learner choice and cross-validation dominate
regression-separate cost. Neural-network training and epoch count commonly
dominate VAEAC cost. Fine batches are unusually expensive for timeseries and
VAEAC, while parallel ARF can require very large amounts of RAM. These
choices belong in the initial resource estimate rather than being treated as
minor tuning details.

## Explore all results

The table below combines every committed approach-level `summary.csv`. Each row
is one configuration aggregated over its replicates. Use the controls above each
column to filter, click headings to sort, and use **Column visibility** to focus
on relevant fields. The first three columns remain fixed while scrolling
horizontally. **CSV** downloads the currently displayed table.

Key timing and memory columns are:

- `bash_wall_median`: median whole-process wall time in seconds;
- `wall_median`: median `explain()` wall time in seconds;
- `wall_iqr`: interquartile range of `explain()` time;
- `ram_mb_median`: median peak RAM in MB;
- `ram_mb_max`: maximum peak RAM across replicates.

<div class="benchmark-table-wide">

```{r benchmark-table}
DT::datatable(
  benchmark_results,
  rownames = FALSE,
  filter = "top",
  extensions = c("Buttons", "FixedColumns"),
  class = "stripe hover compact nowrap",
  width = "100%",
  options = list(
    dom = "Bfrtip",
    buttons = list(
      "copy",
      list(extend = "csv", filename = "shapr-benchmark-summary"),
      "colvis"
    ),
    scrollX = TRUE,
    scrollY = "65vh",
    scrollCollapse = TRUE,
    fixedColumns = list(leftColumns = 3),
    pageLength = 25,
    lengthMenu = list(c(10, 25, 50, 100, -1), c("10", "25", "50", "100", "All")),
    autoWidth = TRUE
  )
) |>
  DT::formatRound(
    columns = c(
      "wall_median",
      "wall_iqr",
      "cpu_user_median",
      "ram_mb_median",
      "ram_mb_max",
      "bash_wall_median",
      "data_load_median",
      "n_iterations_median"
    ),
    digits = 3
  )
```

</div>
