---
title: "`shapr`: Explaining individual machine learning predictions with Shapley values"
author: "Camilla Lingjærde, Martin Jullum & Nikolai Sellereite"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Understanding shapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width =7,
  fig.height = 3
)
```

```{r setup, include=FALSE, warning=FALSE}
library(shapr)
```
 
 
> [Introduction](#intro)

> [Overview of Package](#overview)

> [The Kernel SHAP Method](#KSHAP)

> [Comparison to `shap`](#compare)

> [An Example](#ex)

<a id="intro"></a>

# Introduction
 
 
The `shapr` package implements an extended version of the Kernel SHAP method for approximating Shapley values (@lundberg2017unified), in which dependence between the features is taken into account (@aas2019explaining). Estimation of Shapley values is of interest when attempting to explain complex machine learning models. Of existing work on interpreting individual predictions, Shapley values is regarded to be the only model-agnostic explanation method with a solid theoretical foundation (@lundberg2017unified). Kernel SHAP is a computationally efficient approximation to Shapley values in higher dimensions, but it assumes independent features. @aas2019explaining extend the Kernel SHAP method to handle dependent features, resulting in more accurate approximations to the true Shapley values. [See the paper for further details](https://arxiv.org/abs/1903.10464).


<a id="overview"></a>

# Overview of Package

## Functions

Here is an overview of the main functions. You can read their documentation and see examples with `?function_name`.

----------------------------- ------------------------------------------------------
Function Name                 Description
----------------------------- ------------------------------------------------------
`compute_kshap`               Computes kernel SHAP values for test data.

`plot_kshap`                  Plots the individual prediction explanations. Uses facet_wrap of ggplot. 

`prepare_kshap`               Get Shapley weights for test data.
------------------- --------------------------------------------------

: Main functions in the `shapr` package.



<a id="KSHAP"></a>

# The Kernel SHAP Method

Assume a predictive model $f(\boldsymbol{x})$ for a response value $y$ with features $\boldsymbol{x}\in \mathbb{R}^M$, trained on a training set, and that we want to explain the predictions for new sets of features. This may be done using ideas from cooperative game theory, letting a single prediction take the place of the game being played and the features the place of the players. Letting $N$ denote the set of all $M$ players, and $S \subseteq N$ be a subset of $|S|$ players, the "contribution" function $v(S)$ describes the total expected sum of payoffs the members of $S$ can obtain by cooperation. The Shapley value (@Shapley53) is one way to distribute the total gains to the players, assuming that they all collaborate. The amount that player $i$ gets is then

$$\phi_i(v) = \phi_i = \sum_{S \subseteq N \setminus\{i\}} \frac{|S| ! (M-| S| - 1)!}{M!}(v(S\cup \{i\})-v(S)),$$


that is, a weighted mean over all subsets $S$ of players not containing player $i$. @lundberg2017unified define the contribution function for a certain subset $S$ of these features $\boldsymbol{x}_S$ as $v(S) = \mbox{E}[f(\boldsymbol{x})|\boldsymbol{x}_S]$, the expected output of the predictive model conditional on the feature values of the subset. @lundberg2017unified names this type of Shapley values SHAP (SHapley Additive exPlanation) values. Since the conditional expectations can be written as 

\begin{equation}
\label{eq:CondExp}
E[f(\boldsymbol{x})|\boldsymbol{x}_s=\boldsymbol{x}_S^*] = E[f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S)|\boldsymbol{x}_S=\boldsymbol{x}_S^*] = \int f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S^*)\,p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)d\boldsymbol{x}_{\bar{S}},
\end{equation}

the conditional distributions $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ are needed to compute the contributions. The Kernel SHAP method of @lundberg2017unified assumes feature independence, so that $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)=p(\boldsymbol{x}_{\bar{S}})$. 
If samples $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$, from $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ are available, the conditional expectation in above can be approximated by 

\begin{equation}
  v_{\text{KerSHAP}}(S) = \frac{1}{K}\sum_{k=1}^K f(\boldsymbol{x}_{\bar{S}}^{k},\boldsymbol{x}_S^*).
\end{equation}

In Kernel SHAP, $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$ are sampled from the $\bar{S}$-part of the training data, *independently* of $\boldsymbol{x}_{S}$. This is motivated by using the training set as the empirical distribution of $\boldsymbol{x}_{\bar{S}}$, and assuming that $\boldsymbol{x}_{\bar{S}}$ is independent of $\boldsymbol{x}_S=\boldsymbol{x}_S^*$.
Due to the indpendencde assumption, if the features in a given model are highly dependent, the Kernel SHAP method may give a completely wrong answer. This can be avoided by estimating the conditional distribution $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ directly and 
generating samples from this distribution. With this small change, the contributions and Shapley values may then be approximated as in the ordinary Kernel SHAP framework. @aas2019explaining propose three different approaches for estimating the conditional probabilities. The methods may also be combined, such that e.g. one method is used when conditioning on a small number of features, while another method is used otherwise.


## Multivariate Gaussian Distribution Approach

The first approach arises from the assumption that the feature vector $\boldsymbol{x}$ stems from a multivariate Gaussian distribution with some mean vector $\boldsymbol{\mu}$ and covariance matrix $\boldsymbol{\Sigma}$. Under this assumption, the conditional distribution $p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$ is also multivariate Gaussian  $\text{N}_{|\bar{\mathcal{S}}|}(\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}},\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}})$, with analytical expressions for the conditional mean vector $\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ and covariance matrix $\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$, see @aas2019explaining for details.
Hence, instead of sampling from the marginal empirical distribution of $\boldsymbol{x}_{\bar{\mathcal{S}}}$ approximated by the training data, we can sample from the Gaussian conditional distribution, which is fitted using the training data. Using the resulting samples  $\boldsymbol{x}_{\bar{\mathcal{S}}}^k, k=1,\ldots,K$, the conditional expectations be approximated as in the Kernel SHAP. 


## Gaussian Copula Approach

If the features are far from multivariate Gaussian, an alternative approach is to instead represent the marginals by their empirical distributions, and model the dependence structure by a Gaussian copula. Assuming a Gaussian copula, we may convert the marginals of the training data to Gaussian features using their empirical distributions, and then fit a multivariate Gaussian distribution to these. 

To produce samples from the conditional distribution $p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$, we convert the marginals of $\boldsymbol{x}_{\mathcal{S}}$ to Gaussians, sample from the conditional Gaussian distribution as above, and convert the marginals of thesamples back to the original distribution. Those samples are then used to approximate the 

sample from the resulting multivariate Gaussian conditional distribution. While other copulas may be used, the Gaussian copula has the benefit that we may use the analytical expressions for the conditionals $\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ and $\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$. Finally, we may convert the marginals back to their original distribution, and use the resulting samples to approximate the conditional expectations as in the Kernel SHAP.  

## Empirical Conditional Distribution Approach

If both the dependence structure and the marginal distributions of $\boldsymbol{x}$ are very far from the Gaussian, neither of the two aforementioned methods will work very well. Few methods exists for the non-parametric estimation of conditional densities, and the classic kernel estimator (@rosenblatt1956) for non-parametric density estimation suffers greatly from the curse of dimensionality and does not provide a way to generate samples from the estimated distribution. For such situations, @aas2019explaining propose an empirical conditional approach to sample approximately from $p(\boldsymbol{x}_{\bar{\mathcal{S}}}|\boldsymbol{x}_{\mathcal{S}}^*)$. The idea is to compute weights $w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^i),\ i=1,...,n_{\text{train}}$ for all training instances based on their Mahalanobis distances (in the $S$ subset only) to the instance $\boldsymbol{x}^*$ to be explained. Instead of sampling from this weighted (conditional) empirical distribution, @aas2019explaining suggests a more efficient varient, using only the $K$ instances with the largest weights:

$$v_{\text{condKerSHAP}}(\mathcal{S}) = \frac{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^{[k]}) f(\boldsymbol{x}_{\bar{\mathcal{S}}}^{[k]},\boldsymbol{x}_{\mathcal{S}}^*)}{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^{[k]})},$$ 

The number of samples $K$ to be used in the approximate prediction can for instance be chosen such that the $K$ largest weights accounts for a fraction $\eta$, for example $0.9$, of the total weight. If $K$ exceeds a certain limit, for instance $5,000$, it might be set to that limit. A bandwidth parameter $\sigma$ used to scale the weights, must also be specified. This choice may be viewed as a bias-variance trade-off. A small $\sigma$ puts most of the weight to a few of the closest training observations and thereby gives low bias, but high variance. When $\sigma \rightarrow \infty$, this method converges to the original Kernel SHAP assuming feature independence. Typically, when the features are highly dependent, a small $\sigma$ is typically needed such that the bias does not dominate. @aas2019explaining show that a proper criterion for selecting $\sigma$ is a small-sample-size corrected version of the AIC known as AICc. As calculation of it is computationally intensive, an approximate version of the selection criterion is also suggested. Details on this is found in @aas2019explaining.



<a id="compare"></a>

# Comparison to `shap`







<a id="ex"></a>

# An Example

`shapr` supports computation of Shapley values with any predictive model which takes a set of numeric features and produces a numeric outcome. 

The following example shows how a simple `xgboost` model is trained using the Boston Housing Data, and how `shapr` can be used to explain the individual predictions. Note that empirical conditional distribution approach is the default, and that the Gaussian and Gaussian copula approaches can be used instead by setting the argument `cond_approach` to either 'Gaussian' or 'copula'.   

```{r,warning=FALSE}
library(MASS)
library(xgboost)
library(shapr)
library(data.table)
library(ggplot2)

data("Boston")

x_var <- c("lstat", "rm", "dis", "indus")
y_var <- "medv"

x_train <- as.matrix(Boston[-(1:6), x_var])
y_train <- Boston[-(1:6), y_var]
x_test <- as.matrix(Boston[1:6, x_var])

# Fitting a basic xgboost model to the training data
model <- xgboost(
  data = x_train,
  label = y_train,
  nround = 20,
  verbose = F
)

# Prepare the data for explanation
l <- prepare_kshap(
  Xtrain = x_train,
  Xtest = x_test
)

# Specifying the phi_0, i.e. the expected prediction without any features
pred_zero <- mean(y_train)

# Computing the actual Shapley values with kernelSHAP accounting for feature dependence using
# the empirical (conditional) distribution approach with bandwidth parameter sigma = 0.1 (default)
explanation <- compute_kshap(
  model = model,
  l = l,
  pred_zero = pred_zero
)

# Printing the Shapley values for the test data
explanation$Kshap

# Plot the resulting explanations for observations 1 and 6, excluding the no-covariate effect
plot_kshap(explanation, l, plot_phi0 = F, plot_which_Xtest = c(1,6))
``` 

The Gaussian approach is used as follows:

```{r}
# Use the Gaussian approach
explanation_Gauss <- compute_kshap(
  model = model,
  l = l,
  cond_approach = 'Gaussian',
  pred_zero = pred_zero
)

# Plot the resulting explanations for observations 1 and 6, excluding the no-covariate effect
plot_kshap(explanation_Gauss, l, plot_phi0 = F, plot_which_Xtest = c(1,6))
```

The Gaussian copula approach is used as follows:

```{r}
# Use the Gaussian copula approach
explanation_copula <- compute_kshap(
  model = model,
  l = l,
  cond_approach = 'copula',
  pred_zero = pred_zero
)

# Plot the resulting explanations for observations 1 and 6, excluding the no-covariate effect
plot_kshap(explanation_copula, l, plot_phi0 = F,plot_which_Xtest = c(1,6))

```



# References

