---
title: "`shapr`: Explaining individual machine learning predictions with Shapley values"
author: "Camilla Lingjærde, Martin Jullum & Nikolai Sellereite"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{`shapr`: Explaining individual machine learning predictions with Shapley values}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3,
  warning = FALSE,
  message = FALSE
)
```

```{r setup, include=FALSE, warning=FALSE}
library(shapr)
```

> [Introduction](#intro)

> [Overview of Package](#overview)

> [The Kernel SHAP Method](#KSHAP)

> [Examples](#ex)

> [Advanced usage](#advanced)

> [Scalability and efficency](#scalability)

> [Comparison to Lundberg & Lee's implementation](#compare)

<a id="intro"></a>

# Introduction

The `shapr` package implements an extended version of the Kernel SHAP
method for approximating Shapley values (@lundberg2017unified), in which
dependence between the features is taken into account
(@aas2019explaining). Estimation of Shapley values is of interest when
attempting to explain complex machine learning models. Of existing work
on interpreting individual predictions, Shapley values is regarded to be
the only model-agnostic explanation method with a solid theoretical
foundation (@lundberg2017unified). Kernel SHAP is a computationally
efficient approximation to Shapley values in higher dimensions, but it
assumes independent features. @aas2019explaining extend the Kernel SHAP
method to handle dependent features, resulting in more accurate
approximations to the true Shapley values. See the
[paper](https://www.sciencedirect.com/sdfe/reader/pii/S0004370221000539/pdf)
(@aas2019explaining) for further details.

<a id="overview"></a>

<br>

# Overview of Package

## Functions

Here is an overview of the main functions. You can read their
documentation and see examples with `?function_name`.

| Function Name      | Description                                                                               |
|:---------------------|:-------------------------------------------------|
| `explain`          | Computes kernel SHAP values for test data.                                                |
| `explain_forecast` | Analogous to `explain`, but for explaining forecasts from time series models.             |
| `plot.shapr`       | Plots the individual prediction explanations. Uses the `ggplot` and `ggbeeswarm` package. |

: Main functions in the `shapr` package.

<a id="KSHAP"></a>

<br>

# The Kernel SHAP Method

Assume a predictive model $f(\boldsymbol{x})$ for a response value $y$
with features $\boldsymbol{x}\in \mathbb{R}^M$, trained on a training
set, and that we want to explain the predictions for new sets of data.
This may be done using ideas from cooperative game theory, letting a
single prediction take the place of the game being played and the
features the place of the players. Letting $N$ denote the set of all $M$
players, and $S \subseteq N$ be a subset of $|S|$ players, the
"contribution" function $v(S)$ describes the total expected sum of
payoffs the members of $S$ can obtain by cooperation. The Shapley value
(@Shapley53) is one way to distribute the total gains to the players,
assuming that they all collaborate. The amount that player $i$ gets is
then

$$\phi_i(v) = \phi_i = \sum_{S \subseteq N \setminus\{i\}} \frac{|S| ! (M-| S| - 1)!}{M!}(v(S\cup \{i\})-v(S)),$$

that is, a weighted mean over all subsets $S$ of players not containing
player $i$. @lundberg2017unified define the contribution function for a
certain subset $S$ of these features $\boldsymbol{x}_S$ as
$v(S) = \mbox{E}[f(\boldsymbol{x})|\boldsymbol{x}_S]$, the expected
output of the predictive model conditional on the feature values of the
subset. @lundberg2017unified names this type of Shapley values SHAP
(SHapley Additive exPlanation) values. Since the conditional
expectations can be written as

```{=tex}
\begin{equation}
\label{eq:CondExp}
E[f(\boldsymbol{x})|\boldsymbol{x}_s=\boldsymbol{x}_S^*] = E[f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S)|\boldsymbol{x}_S=\boldsymbol{x}_S^*] = 
\int f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S^*)\,p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)d\boldsymbol{x}_{\bar{S}},
\end{equation}
```
the conditional distributions
$p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ are
needed to compute the contributions. The Kernel SHAP method of
@lundberg2017unified assumes feature independence, so that
$p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)=p(\boldsymbol{x}_{\bar{S}})$.
If samples $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$, from
$p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ are
available, the conditional expectation in above can be approximated by

```{=tex}
\begin{equation}
  v_{\text{KerSHAP}}(S) = \frac{1}{K}\sum_{k=1}^K f(\boldsymbol{x}_{\bar{S}}^{k},\boldsymbol{x}_S^*).
\end{equation}
```
In Kernel SHAP, $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$ are sampled
from the $\bar{S}$-part of the training data, *independently* of
$\boldsymbol{x}_{S}$. This is motivated by using the training set as the
empirical distribution of $\boldsymbol{x}_{\bar{S}}$, and assuming that
$\boldsymbol{x}_{\bar{S}}$ is independent of
$\boldsymbol{x}_S=\boldsymbol{x}_S^*$. Due to the independence
assumption, if the features in a given model are highly dependent, the
Kernel SHAP method may give a completely wrong answer. This can be
avoided by estimating the conditional distribution
$p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$
directly and generating samples from this distribution. With this small
change, the contributions and Shapley values may then be approximated as
in the ordinary Kernel SHAP framework. @aas2019explaining propose three
different approaches for estimating the conditional probabilities which
are implemented: `empirical`, `gaussian` and `copula`. The package also
implements the method `ctree` method from @redelmeier2020explaining. The
original `independence` approach of @lundberg2017unified is also
available. The methods may also be combined, such that e.g. one method
is used when conditioning on a small number of features, while another
method is used otherwise.

<a id="gaussian"></a>

## Multivariate Gaussian Distribution Approach

The first approach arises from the assumption that the feature vector
$\boldsymbol{x}$ stems from a multivariate Gaussian distribution with
some mean vector $\boldsymbol{\mu}$ and covariance matrix
$\boldsymbol{\Sigma}$. Under this assumption, the conditional
distribution
$p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$
is also multivariate Gaussian\
$\text{N}_{|\bar{\mathcal{S}}|}(\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}},\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}})$,
with analytical expressions for the conditional mean vector
$\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ and covariance matrix
$\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$, see
@aas2019explaining for details. Hence, instead of sampling from the
marginal empirical distribution of $\boldsymbol{x}_{\bar{\mathcal{S}}}$
approximated by the training data, we can sample from the Gaussian
conditional distribution, which is fitted using the training data. Using
the resulting samples
$\boldsymbol{x}_{\bar{\mathcal{S}}}^k, k=1,\ldots,K$, the conditional
expectations be approximated as in the Kernel SHAP.

<a id="copula"></a>

## Gaussian Copula Approach

If the features are far from multivariate Gaussian, an alternative
approach is to instead represent the marginals by their empirical
distributions, and model the dependence structure by a Gaussian copula.
Assuming a Gaussian copula, we may convert the marginals of the training
data to Gaussian features using their empirical distributions, and then
fit a multivariate Gaussian distribution to these.

To produce samples from the conditional distribution
$p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$,
we convert the marginals of $\boldsymbol{x}_{\mathcal{S}}$ to Gaussians,
sample from the conditional Gaussian distribution as above, and convert
the marginals of the samples back to the original distribution. Those
samples are then used to approximate the sample from the resulting
multivariate Gaussian conditional distribution. While other copulas may
be used, the Gaussian copula has the benefit that we may use the
analytical expressions for the conditionals
$\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ and
$\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$. Finally, we may
convert the marginals back to their original distribution, and use the
resulting samples to approximate the conditional expectations as in the
Kernel SHAP.

<a id="empirical"></a>

## Empirical Conditional Distribution Approach

If both the dependence structure and the marginal distributions of
$\boldsymbol{x}$ are very far from the Gaussian, neither of the two
aforementioned methods will work very well. Few methods exists for the
non-parametric estimation of conditional densities, and the classic
kernel estimator (@rosenblatt1956) for non-parametric density estimation
suffers greatly from the curse of dimensionality and does not provide a
way to generate samples from the estimated distribution. For such
situations, @aas2019explaining propose an empirical conditional approach
to sample approximately from
$p(\boldsymbol{x}_{\bar{\mathcal{S}}}|\boldsymbol{x}_{\mathcal{S}}^*)$.
The idea is to compute weights
$w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^i),\ i=1,...,n_{\text{train}}$
for all training instances based on their Mahalanobis distances (in the
$S$ subset only) to the instance $\boldsymbol{x}^*$ to be explained.
Instead of sampling from this weighted (conditional) empirical
distribution, @aas2019explaining suggests a more efficient variant,
using only the $K$ instances with the largest weights:

$$v_{\text{condKerSHAP}}(\mathcal{S}) = \frac{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,
\boldsymbol{x}^{[k]}) f(\boldsymbol{x}_{\bar{\mathcal{S}}}^{[k]},
\boldsymbol{x}_{\mathcal{S}}^*)}{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^{[k]})},$$

The number of samples $K$ to be used in the approximate prediction can
for instance be chosen such that the $K$ largest weights accounts for a
fraction $\eta$, for example $0.9$, of the total weight. If $K$ exceeds
a certain limit, for instance $5,000$, it might be set to that limit. A
bandwidth parameter $\sigma$ used to scale the weights, must also be
specified. This choice may be viewed as a bias-variance trade-off. A
small $\sigma$ puts most of the weight to a few of the closest training
observations and thereby gives low bias, but high variance. When
$\sigma \rightarrow \infty$, this method converges to the original
Kernel SHAP assuming feature independence. Typically, when the features
are highly dependent, a small $\sigma$ is typically needed such that the
bias does not dominate. @aas2019explaining show that a proper criterion
for selecting $\sigma$ is a small-sample-size corrected version of the
AIC known as AICc. As calculation of it is computationally intensive, an
approximate version of the selection criterion is also suggested.
Details on this is found in @aas2019explaining.

<a id="ex"></a>

<br>

## Conditional Inference Tree Approach

The previous three methods can only handle numerical data. This means
that if the data contains categorical/discrete/ordinal features, the
features first have to be one-hot encoded. When the number of
levels/features is large, this is not feasible. An approach that handles
mixed (i.e numerical, categorical, discrete, ordinal) features and both
univariate and multivariate responses is conditional inference trees
(@hothorn2006unbiased).

Conditional inference trees is a special tree fitting procedure that
relies on hypothesis tests to choose both the splitting feature and the
splitting point. The tree fitting procedure is sequential: first a
splitting feature is chosen (the feature that is least independent of
the response), and then a splitting point is chosen for this feature.
This decreases the chance of being biased towards features with many
splits (@hothorn2006unbiased).

We use conditional inference trees (*ctree*) to model the conditional
distribution,
$p(\boldsymbol{x}_{\bar{\mathcal{S}}}|\boldsymbol{x}_{\mathcal{S}}^*)$,
found in the Shapley methodology. First, we fit a different conditional
inference tree to each conditional distribution. Once a tree is fit for
given dependent features, the end node of
$\boldsymbol{x}_{\mathcal{S}}^*$ is found. Then, we sample from this end
node and use the resulting samples,
$\boldsymbol{x}_{\bar{\mathcal{S}}}^k, k=1,\ldots,K$, when approximating
the conditional expectations as in Kernel SHAP. See
@redelmeier2020explaining for more details.

The conditional inference trees are fit using the *party* and *partykit*
packages (@partykit_package).

## Categorical Approach

When the features are all categorical, we can estimate the conditional
expectations using basic statistical formulas. For example, if we have
three features, $x_1, x_2, x_3$ with three levels each (indicated as 1,
2, 3), and we are provided with a table of counts indicating how many
times each combination of feature values occurs, we can estimate the
marginal and conditional probabilities as follows. Marginal
probabilities are estimated by dividing the number of times a given
feature (or features) takes on a certain value in the data set with the
total number of observations in the data set. Condititional
probabilities (for example, $P(X_1 = 1 | X_2 = 1)$) are estimated by
first subsetting the data set to reflect the conditioning (i.e.,
extracting all rows where $X_2 = 1$), and then dividing the number of
times the feature on the left hand side of $|$ takes the given value in
this subset by the total number of observations in this subset. Once the
marginal and conditional probabilities are estimated for all
combinations of feature values, each conditional expectation can be
calculated. For example, the expected value of $X_1$ given $X_2 = 1$ and
$X_3 = 2$ is
$$E(X_1|X_2, X_3) = \sum_{x}x P(X_1 = x | X_2=1, X_3=2) = \sum_{x} x \frac{P(X_1 = x, X_2 = 1, X_3 = 2)}{P(X_2=1, X_3=2)}.$$.

<a id="ex"></a>

<br>

# Examples

`shapr` supports computation of Shapley values with any predictive model
which takes a set of numeric features and produces a numeric outcome.
Note that the ctree method takes both numeric and categorical variables.
Check under "Advanced usage" for an example of how this can be done.

The following example shows how a simple `xgboost` model is trained
using the `airquality` dataset, and how `shapr` can be used to explain
the individual predictions. Note that the empirical conditional
distribution approach is the default (i.e. `approach = "empirical"`).
The Gaussian, Gaussian copula, ctree or independence approaches can be
used instead by setting the argument `approach` to either `"gaussian"`,
`"copula"`, `"ctree"`, `"categorical"` or `"independence"` in the code
below.

```{r, warning=FALSE}
library(xgboost)
library(data.table)

data("airquality")
data <- data.table::as.data.table(airquality)
data <- data[complete.cases(data), ]

x_var <- c("Solar.R", "Wind", "Temp", "Month")
y_var <- "Ozone"

ind_x_explain <- 1:6
x_train <- data[-ind_x_explain, ..x_var]
y_train <- data[-ind_x_explain, get(y_var)]
x_explain <- data[ind_x_explain, ..x_var]

# Fitting a basic xgboost model to the training data
model <- xgboost::xgboost(
  data = as.matrix(x_train),
  label = y_train,
  nround = 20,
  verbose = FALSE
)

# Specifying the phi_0, i.e. the expected prediction without any features
p0 <- mean(y_train)

# Computing the actual Shapley values with kernelSHAP accounting for feature dependence using
# the empirical (conditional) distribution approach with bandwidth parameter sigma = 0.1 (default)
explanation <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0
)

# Printing the Shapley values for the test data.
# For more information about the interpretation of the values in the table, see ?shapr::explain.
print(explanation$shapley_values)

# Plot the resulting explanations for observations 1 and 6
plot(explanation, bar_plot_phi0 = FALSE, index_x_explain = c(1, 6))
```

There are multiple plot options specified by the `plot_type` argument in
`plot`. The `waterfall` option shows the changes in the prediction score
due to each features contribution (their Shapley values):

There are multiple plot options specified by the `plot_type` argument in `plot`. 
The `waterfall` option shows the changes in the prediction score due to each features contribution (their Shapley values): 

```{r}
plot(explanation, plot_type = "waterfall", index_x_explain = c(1, 6))
```

The other two plot options, `"beeswarm"` and `"scatter"`, can be useful
when you have many observations that you want to explain. For the
purpose of illustration, we explain the whole `airquality` dataset
(including the training data) for these plot types. The
`plot_type = "beeswarm"` summarises the distribution of the Shapley
values along the x-axis across all features. Each point gives the
Shapley value of a given instance, where the points are colored by the
feature value of that instance:

```{r}
x_explain_many <- data[, ..x_var]
explanation_plot <- explain(
  model = model,
  x_explain = x_explain_many,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0
)
plot(explanation_plot, plot_type = "beeswarm")
```

The `plot_type = "scatter"` plots the feature values on the x-axis and
Shapley values on the y-axis, as well as (optionally) a background
scatter_hist showing the distribution of the feature data:

```{r}
plot(explanation_plot, plot_type = "scatter", scatter_hist = TRUE)
```

We can use mixed (i.e continuous, categorical, ordinal) data with ctree.
Use ctree with mixed data in the following manner:

```{r}
# convert the month variable to a factor
data[, Month_factor := as.factor(Month)]

data_train_cat <- data[-ind_x_explain, ]
data_explain_cat <- data[ind_x_explain, ]

x_var_cat <- c("Solar.R", "Wind", "Temp", "Month_factor")

x_train_cat <- data_train_cat[, ..x_var_cat]
x_explain_cat <- data_explain_cat[, ..x_var_cat]

# Fitting an lm model here as xgboost does not handle categorical features directly
# (work around in example below)
lm_formula <- as.formula(paste0(y_var, " ~ ", paste0(x_var_cat, collapse = " + ")))

model_lm_cat <- lm(lm_formula, data_train_cat)

p0 <- mean(y_train)
explanation_lm_cat <- explain(
  model = model_lm_cat,
  x_explain = x_explain_cat,
  x_train = x_train_cat,
  approach = "ctree",
  prediction_zero = p0
)

# Plot the resulting explanations for observations 1 and 6, excluding
# the no-covariate effect
plot(explanation_lm_cat, bar_plot_phi0 = FALSE, index_x_explain = c(1, 6))
```

We can specify parameters used to build the conditional inference trees
in the following manner. Default values are based on
@hothorn2006unbiased.

```{r}
# Use the conditional inference tree approach
# We can specify parameters used to building trees by specifying mincriterion,
# minsplit, minbucket
explanation_ctree <- explain(
  model = model_lm_cat,
  x_explain = x_explain_cat,
  x_train = x_train_cat,
  approach = "ctree",
  prediction_zero = p0,
  ctree.mincriterion = 0.80,
  ctree.minsplit = 20,
  ctree.minbucket = 20
)
# Default parameters (based on (Hothorn, 2006)) are:
# mincriterion = 0.95
# minsplit = 20
# minbucket = 7
```

If **all** features are categorical, one may use the categorical
approach as follows:

```{r}
# For the sake of illustration, convert ALL features to factors
data[, Solar.R_factor := as.factor(cut(Solar.R, 10))]
data[, Wind_factor := as.factor(cut(Wind, 3))]
data[, Temp_factor := as.factor(cut(Temp, 2))]
data[, Month_factor := as.factor(Month)]

data_train_all_cat <- data[-ind_x_explain, ]
data_explain_all_cat <- data[ind_x_explain, ]


x_var_all_cat <- c("Solar.R_factor", "Wind_factor", "Temp_factor", "Month_factor")

x_train_all_cat <- data_train_all_cat[, ..x_var_all_cat]
x_explain_all_cat <- data_explain_all_cat[, ..x_var_all_cat]

# Fit an lm model here
lm_formula_all_cat <- as.formula(paste0(y_var, " ~ ", paste0(x_var_all_cat, collapse = " + ")))

model_lm_all_cat <- lm(lm_formula_all_cat, data_train_all_cat)

explanation_cat_method <- explain(
  model = model_lm_all_cat,
  x_explain = x_explain_all_cat,
  x_train = x_train_all_cat,
  approach = "categorical",
  prediction_zero = p0
)
```

Shapley values can be used to explain any predictive model. For
predictive models taking time series as input, `approach='timeseries'`
can be used. In such models, joint behavior of consecutive time points
is often more important for the outcome than the single time points.
Therefore, it makes sense to derive Shapley value segments of the time
series instead of for each single time point. In `shapr` this can be
achieved through the `group` attribute. Other optional parameters of
`approach='timeseries'` are `timeseries.fixed_sigma_vec` and
`timeseries.bounds` (a vector indicating upper and lower bounds of the
time series if necessary).

```{r}
# Simulate time series data with AR(1)-structure
set.seed(1)
data_ts <- data.frame(matrix(NA, ncol = 41, nrow = 4))
for (n in 1:100) {
  set.seed(n)
  e <- rnorm(42, mean = 0, sd = 1)

  m_1 <- 0
  for (i in 2:length(e)) {
    m_1[i] <- 1 + 0.8 * m_1[i - 1] + e[i]
  }
  data_ts[n, ] <- m_1[-1]
}
data_ts <- data.table::as.data.table(data_ts)

x_var_ts <- paste0("X", 1:40)
y_var_ts <- "X41"

ind_x_explain <- 1:6
data_ts_train <- data_ts[-ind_x_explain]

# Creating a predictive model (for illustration just predicting the next point in the time series with a linear model)
lm_ts_formula <- as.formula(X41 ~ .)
model_lm_ts <- lm(lm_ts_formula, data_ts_train)

x_explain_ts <- data_ts[ind_x_explain, ..x_var_ts]
x_train_ts <- data_ts[-ind_x_explain, ..x_var_ts]

# Spitting the time series into 4 segments
group_ts <- list(
  S1 = paste0("X", 1:10),
  S2 = paste0("X", 11:20),
  S3 = paste0("X", 21:30),
  S4 = paste0("X", 31:40)
)


p0_ts <- mean(unlist(data_ts_train[, ..y_var_ts]))

explanation_timeseries <- explain(
  model = model_lm_ts,
  x_explain = x_explain_ts,
  x_train = x_train_ts,
  approach = "timeseries",
  prediction_zero = p0_ts,
  group = group_ts
)
```

## Main arguments in `explain`

When using `explain`, the default behavior is to use all feature
combinations in the Shapley formula. Kernel SHAP's sampling based
approach may be used by specifying `n_combinations`, which is the number
of unique feature combinations to sample. If not specified, the exact
method is used. The computation time grows approximately exponentially
with the number of features. The training data and the model whose
predictions we wish to explain must be provided through the arguments
`x_train` and `model`. The data whose predicted values we wish to
explain must be given by the argument `x_explain`. Note that both
`x_train` and `x_explain` must be a `data.frame` or a `matrix`, and all
elements must be finite numerical values. Currently we do not support
missing values. The default approach when computing the Shapley values
is the empirical approach (i.e. `approach = "empirical"`). If you'd like
to use a different approach you'll need to set `approach` equal to
either `copula` or `gaussian`, or a vector of them, with length equal to
the number of features. If a vector, a combined approach is used, and
element `i` indicates the approach to use when conditioning on `i`
variables. For more details see [Combined approach](#combined) below.

When computing the kernel SHAP values by `explain`, the maximum number
of samples to use in the Monte Carlo integration for every conditional
expectation is controlled by the argument `n_samples` (default equals
`1000`). The computation time grows approximately linear with this
number. You will also need to pass a numeric value for the argument
`prediction_zero`, which represents the prediction value when not
conditioning on any features. We recommend setting this equal to the
mean of the response, but other values, like the mean prediction of a
large test data set is also a possibility. If the empirical method is
used, specific settings for that approach, like a vector of fixed
$\sigma$ values can be specified through the argument
`empirical.fixed_sigma`. See `?explain` for more information. If
`approach = "gaussian"`, you may specify the mean vector and covariance
matrix of the data generating distribution by the arguments
`gaussian.mu` and `gaussian.cov_mat`. If not specified, they are
estimated from the training data.

## Explaining a forecasting model using `explain_forecast`

`shapr` provides a specific function, `explain_forecast`, to explain
forecasts from time series models, at one or more steps into the future.
The main difference compared to `explain` is that the data is supplied
as (set of) time series, in addition to index arguments (`train_idx` and
`explain_idx`) specifying which time points that represents the train
and explain parts of the data. See `?explain_forecast` for more
information.

To demonstrate how to use the function, 500 observations are generated
which follow an AR(1) structure, i.e.
$y_t = 0.5 y_{t-1} + \varepsilon_t$. To this data an arima model of
order (2, 0, 0) is fitted, and we therefore would like to explain the
forecasts in terms of the two previous lags of the time series. This is
is specified through the argument `explain_y_lags = 2`. Note that some
models may also put restrictions on the amount of data required to make
a forecast. The AR(2) model we used there, for instance, requires two
previous time point to make a forecast.

In the example, two separate forecasts, each three steps ahead, are
explained. To set the starting points of the two forecasts,
`explain_idx` is set to `499:500`. This means that one forecast of
$t = (500, 501, 502)$ and another of $t = (501, 502, 503)$, will be
explained. In other words, `explain_idx` tells `shapr` at which points
in time data was available up until, when making the forecast to
explain.

In the same way, `train_idx` denotes the points in time used to estimate
the conditional expectations used to explain the different forecasts.
Note that since we want to explain the forecasts in terms of the two
previous lags (`explain_y_lags = 2`), the smallest value of `train_idx`
must also be 2, because at time $t = 1$ there was only a single
observation available.

Since the data is stationary, the mean of the data is used as value of
`prediction_zero` (i.e. $\phi_0$). This can however be chosen
differently depending on the data and application.

For a multivariate model such as a VAR (Vector AutoRegressive model), it
may be of more interesting to explain the impact of each variable,
rather than each lag of each variable. This can be done by setting
`group_lags = TRUE`.

```{r}
# Simulate time series data with AR(1)-structure.
set.seed(1)
data_ts <- data.frame(Y = arima.sim(list(order = c(1, 0, 0), ar = .5), n = 500))
data_ts <- data.table::as.data.table(data_ts)

# Fit an ARIMA(2, 0, 0) model.
arima_model <- arima(data_ts, order = c(2, 0, 0))

# Set prediction zero as the mean of the data for each forecast point.
p0_ar <- rep(mean(data_ts$Y), 3)

# Explain forecasts from points t = 499 and t = 500.
explain_idx <- 499:500

explanation_forecast <- explain_forecast(
  model = arima_model,
  y = data_ts,
  train_idx = 2:498,
  explain_idx = 499:500,
  explain_y_lags = 2,
  horizon = 3,
  approach = "empirical",
  prediction_zero = p0_ar,
  group_lags = FALSE
)
explanation_forecast
```

Note that for a multivariate model such as a VAR (Vector AutoRegressive
model), or for models also including several exogenous variables, it may
be of more informative to explain the impact of each variable, rather
than each lag of each variable. This can be done by setting
`group_lags = TRUE`. This does not make sense for this model, however,
as that would result in decomposing the forecast into a single group.

We now give a more hands on example of how to use the `explain_forecast`
function. Say that we have an AR(2) model which describes the change
over time of the variable `Temp` in the dataset `airquality`. It seems
reasonable to assume that the temperature today should affect the
temperature tomorrow. To a lesser extent, we may also suggest that the
temperature today should also have an impact on that of the day after
tomorrow.

We start by building our AR(2) model, naming it `model_ar_temp`. This
model is then used to make a forecast of the temperature of the day that
comes after the last day in the data, this forecast starts from index
153.

```{r}
data <- data.table::as.data.table(airquality)

model_ar_temp <- ar(data$Temp, order = 2)

predict(model_ar_temp, n.ahead = 2)$pred
```

First, we pass the model and the data as `model` and `y`. Since we have
an AR(2) model, we want to explain the forecasts in terms of the two
previous lags, whihc we specify with `explain_y_lags = 2`. Then, we let
`shapr` know which time indices to use as training data through the
argument `train_idx`. We use `2:152`, meaning that we skip the first
index, as we want to explain the two previous lags. Letting the training
indices go up until 152 means that every point in time except the first
and last will be used as training data.

The last index, 153 is passed as the argument `explain_idx`, which means
that we want to explain a forecast made from time point 153 in the data.
The argument `horizon` is set to 2 in order to explain a forecast of
length 2.

The argument `prediction_zero` is set to the mean of the time series,
and is repeated two times. Each value of `prediction_zero` is the
baseline for each forecast horizon. In our example, we assume that given
no effect from the two lags, the temperature would just be the average
during the observed period. Finally, we opt to not group the lags by
setting `group_lags` to `FALSE`. This means that lag 1 and 2 will be
explained separately. Grouping lags may be more interesting to do in a
model with multiple variables, as it is then possible to explain each
variable separately.

```{r}
explanation <- explain_forecast(
  model = model_ar_temp,
  y = data[, "Temp"],
  train_idx = 2:152,
  explain_idx = 153,
  explain_y_lags = 2,
  horizon = 2,
  approach = "empirical",
  prediction_zero = rep(mean(data$Temp), 2),
  group_lags = FALSE,
  n_batches = 1,
  timing = FALSE
)

print(explanation)
```

The results are presented per value of `explain_idx` and forecast
horizon. We can see that the mean temperature was around 77.9 degrees.
At horizon 1, the first lag in the model caused it to be 6.6 degrees
lower, and the second lag had just a minor effect. At horizon 2, the
first lag has a slightly smaller negative impact, and the second lag has
a slightly larger impact.

It is also possible to explain a forecasting model which uses exogenous
regressors. The previous example is expanded to use an ARIMA(2,0,0)
model with `Wind` as an exogenous regressor. Since the exogenous
regressor must be available for the predicted time points, the model is
just fit on the 151 first observations, leaving two observations of
`Wind` to be used as exogenous values during the prediction phase.

```{r}
data <- data.table::as.data.table(airquality)

data_fit <- data[seq_len(151), ]

model_arimax_temp <- arima(data_fit$Temp, order = c(2, 0, 0), xreg = data_fit$Wind)

newxreg <- data[-seq_len(151), "Wind", drop = FALSE]

predict(model_arimax_temp, n.ahead = 2, newxreg = newxreg)$pred
```

The `shapr` package can then explain not only the two autoregressive
lags, but also the single lag of the exogenous regressor. In order to do
so, the `Wind` variable is passed as the argument `xreg`, and
`explain_xreg_lags` is set to 1. Notice how only the first 151
observations are used for `y` and all 153 are used for `xreg`. This
makes it possible for `shapr` to not only explain the effect of the
first lag of the exogenous variable, but also the contemporary effect
during the forecasting period.

```{r}
explanation <- explain_forecast(
  model = model_ar_temp,
  y = data_fit[, "Temp"],
  xreg = data[, "Wind"],
  train_idx = 2:150,
  explain_idx = 151,
  explain_y_lags = 2,
  explain_xreg_lags = 1,
  horizon = 2,
  approach = "empirical",
  prediction_zero = rep(mean(data_fit$Temp), 2),
  group_lags = FALSE,
  n_batches = 1,
  timing = FALSE
)

print(explanation$shapley_values)
```

<a id="advanced"></a>

<br>

# Advanced usage

<a id="combined"></a>

## Combined approach

In addition to letting the user select one of the five aforementioned
approaches for estimating the conditional distribution of the data (i.e.
`approach` equals either [`"gaussian"`](#gaussian),
[`"copula"`](#copula), [`"empirical"`](#empirical) or
[`"ctree"`](#ctree) or [`"categorical"`](#categorical)), the package
allows the user to combine the given approaches. To simplify the usage,
the flexibility is restricted such that the same approach is used when
conditioning on the same number of features. This is also in line
@aas2019explaining [, Section 3.4].

This can be done by setting `approach` equal to a character vector,
where the length of the vector is equal to the number of features in the
model. Consider a situation where you have trained a model that consists
of 10 features, and you would like to use the `"empirical"` approach
when you condition on 1-3 features, the `"copula"` approach when you
condition on 4-5 features, and the `"gaussian"` approach when
conditioning on 6 or more features. This can be applied by simply
passing
`approach = c(rep("empirical", 3), rep("copula", 2), rep("gaussian", 5))`,
i.e. `approach[i]` determines which method to use when conditioning on
`i` features.

The code below exemplifies this approach for a case where there are four
features, using `"empirical", "copula"` and `"gaussian"` when
conditioning on respectively 1, 2 and 3-4 features. Note that it does
not matter what method that is specified when conditioning on all
features, as that equals the actual prediction regardless of the
specified approach.

```{r}
# Use the combined approach
explanation_combined <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = c("empirical", "copula", "gaussian", "gaussian"),
  prediction_zero = p0
)
# Plot the resulting explanations for observations 1 and 6, excluding
# the no-covariate effect
plot(explanation_combined, bar_plot_phi0 = FALSE, index_x_explain = c(1, 6))
```

As a second example using `"ctree"` for the first 3 features and
`"empirical"` for the last:

```{r}
# Use the combined approach
explanation_combined <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = c("ctree", "ctree", "ctree", "empirical"),
  prediction_zero = p0
)
```

## Explain groups of features

In some cases, especially when the number of features is very large, it
may be more appropriate to explain predictions in terms of groups of
features instead of single features, see (@jullum2021efficient) for
intuition and real world examples. Explaining prediction in terms of
groups of features is very easy using `shapr`:

```{r}
# Define the feature groups
group_list <- list(
  A = c("Temp", "Month"),
  B = c("Wind", "Solar.R")
)

# Use the empirical approach
explanation_group <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  group = group_list
)
# Prints the group-wise explanations
explanation_group
# Plots the group-wise explanations
plot(explanation_group, bar_plot_phi0 = TRUE, index_x_explain = c(1, 6))
```

## Explain custom models

`shapr` currently natively supports explanation of predictions from
models fitted with the following functions:

-   `stats::lm`
-   `stats::glm`
-   `ranger::ranger`
-   `mgcv::gam`
-   `xgboost::xgboost`/`xgboost::xgb.train`

Any continuous response regression model or binary classification model
of these model classes, can be explained with the package directly as
exemplified above. Moreover, essentially any feature dependent
prediction model can be explained by the package by specifying two (or
one) simple additional functions for your model.

*Note: The below procedure for specifying custom models was changed in
shapr v0.3.0* The first function is `predict_model`, taking the model
and data (as a `matrix` or `data.frame/data.table`) as input and
outputting the corresponding prediction as a numeric vector. The second
(optional, but highly recommended) function is `get_model_specs`, taking
the model as input and outputting a list with the following elements:
*labels* (vector with the feature names to compute Shapley values for),
*classes* (a named vector with the labels as names and the class type as
elements), *factor_levels* (a named list with the labels as names and
vectors with the factor levels as elements (NULL if the feature is not a
factor)). The `get_model_specs` function is used to check that the
format of the data passed to `explain` have the correct format in terms
of the necessary feature columns being available and having the correct
class/attributes. It is highly recommended to do such checks in order to
ensure correct usage of `explain`. If, for some reason, such checking is
not desirable, one does not have to provide the `get_model_specs`
function. This will, however, throw a warning that all feature
consistency checking against the model is disabled.

Once the above functions are created, you can explain predictions from
this model as before by passing the functions through the input
arguments `predict_model` and `get_model_specs` of `explain()`.

These functions **can** be made general enough to handle all supported
model types of that class, or they can be made minimal, possibly only
allowing explanation of the specific version of the model class at hand.
Below we give examples of both full support versions of these functions
and a minimal version which skips the `get_model_specs` function. We do
this for the `gbm` model class from the `gbm` package, fitted to the
same airquality data set as used above.

```{r,warning=TRUE, message=TRUE}
library(gbm)

formula_gbm <- as.formula(paste0(y_var, "~", paste0(x_var, collapse = "+")))
# Fitting a gbm model
set.seed(825)
model_gbm <- gbm::gbm(
  formula_gbm,
  data = cbind(x_train, Ozone = y_train),
  distribution = "gaussian"
)

#### Full feature versions of the three required model functions ####
MY_predict_model <- function(x, newdata) {
  if (!requireNamespace("gbm", quietly = TRUE)) {
    stop("The gbm package is required for predicting train models")
  }
  model_type <- ifelse(
    x$distribution$name %in% c("bernoulli", "adaboost"),
    "classification",
    "regression"
  )
  if (model_type == "classification") {
    predict(x, as.data.frame(newdata), type = "response", n.trees = x$n.trees)
  } else {
    predict(x, as.data.frame(newdata), n.trees = x$n.trees)
  }
}
MY_get_model_specs <- function(x) {
  feature_specs <- list()
  feature_specs$labels <- labels(x$Terms)
  m <- length(feature_specs$labels)
  feature_specs$classes <- attr(x$Terms, "dataClasses")[-1]
  feature_specs$factor_levels <- setNames(vector("list", m), feature_specs$labels)
  feature_specs$factor_levels[feature_specs$classes == "factor"] <- NA # model object doesn't contain factor levels info
  return(feature_specs)
}

# Compute the Shapley values
set.seed(123)
p0 <- mean(y_train)
explanation_custom <- explain(
  model = model_gbm,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  predict_model = MY_predict_model,
  get_model_specs = MY_get_model_specs
)

# Plot results
plot(explanation_custom, index_x_explain = c(1, 6))


#### Minimal version of the three required model functions ####
# Note: Working only for this exact version of the model class
# Avoiding to define get_model_specs skips all feature
# consistency checking between your data and model
MY_MINIMAL_predict_model <- function(x, newdata) {
  predict(x, as.data.frame(newdata), n.trees = x$n.trees)
}

# Compute the Shapley values
set.seed(123)
explanation_custom_minimal <- explain(
  model = model_gbm,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  predict_model = MY_MINIMAL_predict_model
)

# Plot results
plot(explanation_custom_minimal, index_x_explain = c(1, 6))
```

<a id="scalability"></a>

<br>

# Scalability and efficency

## Batch computation

The computational complexity of Shapley value based explanations grows
fast in the number of features, as the number of conditional
expectations one needs to estimate in the Shapley formula grows
exponentially. As outlined [above](#KSHAP), the estimating of each of
these conditional expectations is also computationally expensive,
typically requiring estimation of a conditional probability
distribution, followed by Monte Carlo integration. These computations
are not only heavy for the CPU, they also require a lot of memory (RAM),
which typically is a limited resource. By doing the most resource hungry
computations (the computation of v(S)) in sequential batches with
different feature subsets $S$, the memory usage can be significantly
reduces. Such batching comes at the cost of an increase in computation
time, which depends on the number of feature subsets (`n_combinations`),
the number of features, the estimation `approach` and so on. When
calling `shapr::explain()`, we allow the user to set the number of
batches with the argument `n_batches`. The default of this argument is
`NULL`, which uses a (hopefully) reasonable trade-off between
computation speed and memory consumption which depends on
`n_combinations` and `approach`. The memory/computation time trade-off
is most apparent for models with more than say 6-7 features. Below we a
basic example where `n_batches=10`:

```{r}
explanation_batch <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  n_batches = 10
)
```

## Parallelized computation

In addition to reducing the memory consumption, the introduction of the
`n_batch` argument allows computation within each batch to be performed in parallel. 
The parallelization in `shapr::explain()` is handled by the
`future_apply` which builds on the `future` environment. The `future`
package works on all OS, allows the user to decide the parallelization
backend (mutliple R procesess or forking), works directly with hpc
clusters, and also supports progress updates for the parallelized task
(see below).

Note that, since it takes some time to duplicate data into different
processes/machines when running in parallel, it is not always
preferrable to run `shapr::explain()` in parallel, at least not with
many parallel sessions (hereby called **workers**). Parallelizatiob also
increases the memory consumption proportionally, so you want to limit
the number of workers for that reason too. In a future version of
`shapr` we will provide experienced based automatic selection of the
number of workers. In the meanwhile, this is all lef to the user, and we
advice that `n_batches` equals some positive integer multiplied by the
number of workers. Below is a basic example of a parallelization with
two workers.

```{r}
library(future)
future::plan(multisession, workers = 2)

explanation_par <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  n_batches = 10
)

future::plan(sequential) # To return to non-parallel computation
```

## Progress updates

`shapr` provides progress updates of the computation of the Shapley
values through the R-package `progressr`. This gives the user full
control over the visual apperance of the progress updates, and also
intergrates seemlessly with the parallelization framework `future` used
by `shapr` (see above). Note that the progress is updated as the batches
are completed, meaning that if you have choosen `n_batches=1`, you will
not get intermediate updates, while if you set `n_batches=10` you will
get updates on every 10% of the computation.

Progress updates are enabled for the current R-session by running the
command `progressr::handlers(local=TRUE)`, before calling
`shapr::explain()`. To use progress updates for only a single call to
`shapr::explain()`, one can wrap the call using
`progressr::with_progress` as follows:
`progressr::with_progress({ shapr::explain() })` The default appearence
of the progress updates is a basic ASCII-based horizontal progress bar.
Other variants can be chosen by passing different strings to
`progressr::handlers()`, some of which require additional packages. If
you are using Rstudio, the progress can be displayed directly in the gui
with `progressr::handlers('rstudio')` (requires the `rstudioapi`
package). If you are running Windows, you may use the pop-up gui
progress bar `progressr::handlers('handler_winprogressbar')`. A wrapper
for progressbar of the flexible `cli` package is also available
`progressr::handlers('cli')` (requires the `cli` package).

For a full list of all progression handlers and the customization
options available with `progressr`, see the `progressr`
[vignette](https://cran.r-project.org/web/packages/progressr/vignettes/progressr-intro.html).
A full code example of using `progressr` with `shapr` is shown below:

```{r,eval = FALSE}
library(progressr)
progressr::handlers(global = TRUE)
# If no progression handler is specified, the txtprogressbar is used
# Other progression handlers:
# progressr::handlers('rstudio') # requires the 'rstudioapi' package
# progressr::handlers('handler_winprogressbar') # Window only
# progressr::handlers('cli') # requires the 'cli' package
explanation <- explain(
  model = model,
  x_explain = x_explain,
  x_train = x_train,
  approach = "empirical",
  prediction_zero = p0,
  n_batches = 10
)

#| [=================================>----------------------]  60% Estimating v(S)
```

<a id="compare"></a>

<br>

# Comparison to Lundberg & Lee's implementation

As mentioned above, the original (independence assuming) Kernel SHAP
implementation can be approximated by setting a large $\sigma$ value
using our empirical approach. If we specify that the distances to *all*
training observations should be used (i.e. setting
`approach = "empirical"` and `empirical.eta = 1` when using `explain`,
we can approximate the original method arbitrarily well by increasing
$\sigma$. For completeness of the `shapr` package, we have also
implemented a version of the original method, which samples training
observations independently with respect to their distances to test
observations (i.e. without the large-$\sigma$ approximation). This
method is available by using `approach = "independence"` in `explain`.

We have compared the results using these two variants with the original
implementation of @lundberg2017unified, available through the Python
library [`shap`](https://github.com/slundberg/shap). As above, we used
the Boston housing data, trained via `xgboost`. We specify that *all*
training observations should be used when explaining all of the 6 test
observations. To run the individual explanation method in the `shap`
Python library we use the `reticulate` `R`-package, allowing Python code
to run within `R`. As this requires installation of Python package, the
comparison code and results is not included in this vignette, but can be
found
[here](https://github.com/NorskRegnesentral/shapr/blob/master/inst/scripts/compare_shap_python.R).
As indicated by the (commented out) results in the file above both
methods in our `R`-package give (up to numerical approximation error)
identical results to the original implementation in the Python `shap`
library.

<br>

# References
