---
title: "`shapr`: Explaining individual machine learning predictions with Shapley values"
author: "Camilla Lingjærde, Martin Jullum & Nikolai Sellereite"
output: rmarkdown::html_vignette
bibliography: ../inst/REFERENCES.bib
vignette: >
  %\VignetteIndexEntry{Understanding shapr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 3
)
```

```{r setup, include=FALSE, warning=FALSE}
library(shapr)
```
 
 
> [Introduction](#intro)

> [Overview of Package](#overview)

> [The Kernel SHAP Method](#KSHAP)

> [Examples](#ex)

> [Comparison to Lundberg & Lee's implementation](#compare)

<a id="intro"></a>

# Introduction
 
 
The `shapr` package implements an extended version of the Kernel SHAP method for approximating 
Shapley values (@lundberg2017unified), in which dependence between the features is taken into 
account (@aas2019explaining). Estimation of Shapley values is of interest when attempting to 
explain complex machine learning models. Of existing work on interpreting individual predictions, 
Shapley values is regarded to be the only model-agnostic explanation method with a solid theoretical 
foundation (@lundberg2017unified). Kernel SHAP is a computationally efficient approximation to 
Shapley values in higher dimensions, but it assumes independent features. @aas2019explaining extend 
the Kernel SHAP method to handle dependent features, resulting in more accurate approximations to 
the true Shapley values. [See the paper for further details](https://arxiv.org/abs/1903.10464).


<a id="overview"></a>

<br>

# Overview of Package

## Functions

Here is an overview of the main functions. You can read their documentation and see examples 
with `?function_name`.

----------------------------- ------------------------------------------------------
Function Name                 Description
----------------------------- ------------------------------------------------------
`shapr`                       Get Shapley weights for test data.

`explain`                     Computes kernel SHAP values for test data.

`plot.shapr`                  Plots the individual prediction explanations. Uses facet_wrap of ggplot. 
----------------------------- --------------------------------------------------

: Main functions in the `shapr` package.



<a id="KSHAP"></a>

<br>

# The Kernel SHAP Method

Assume a predictive model $f(\boldsymbol{x})$ for a response value $y$ with features 
$\boldsymbol{x}\in \mathbb{R}^M$, trained on a training set, and that we want to explain the 
predictions for new sets of features. This may be done using ideas from cooperative game theory, 
letting a single prediction take the place of the game being played and the features the place of 
the players. Letting $N$ denote the set of all $M$ players, and $S \subseteq N$ be a subset 
of $|S|$ players, the "contribution" function $v(S)$ describes the total expected sum of payoffs 
the members of $S$ can obtain by cooperation. The Shapley value (@Shapley53) is one way to 
distribute the total gains to the players, assuming that they all collaborate. The amount that 
player $i$ gets is then

$$\phi_i(v) = \phi_i = \sum_{S \subseteq N \setminus\{i\}} \frac{|S| ! (M-| S| - 1)!}{M!}(v(S\cup \{i\})-v(S)),$$


that is, a weighted mean over all subsets $S$ of players not containing player $i$. 
@lundberg2017unified define the contribution function for a certain subset $S$ of these features 
$\boldsymbol{x}_S$ as $v(S) = \mbox{E}[f(\boldsymbol{x})|\boldsymbol{x}_S]$, the expected output 
of the predictive model conditional on the feature values of the subset. @lundberg2017unified names
this type of Shapley values SHAP (SHapley Additive exPlanation) values. Since the conditional 
expectations can be written as 

\begin{equation}
\label{eq:CondExp}
E[f(\boldsymbol{x})|\boldsymbol{x}_s=\boldsymbol{x}_S^*] = E[f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S)|\boldsymbol{x}_S=\boldsymbol{x}_S^*] = \int f(\boldsymbol{x}_{\bar{S}},\boldsymbol{x}_S^*)\,p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)d\boldsymbol{x}_{\bar{S}},
\end{equation}

the conditional distributions $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ are 
needed to compute the contributions. The Kernel SHAP method of @lundberg2017unified assumes feature 
independence, so that $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)=p(\boldsymbol{x}_{\bar{S}})$. 
If samples $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$, from $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ 
are available, the conditional expectation in above can be approximated by 

\begin{equation}
  v_{\text{KerSHAP}}(S) = \frac{1}{K}\sum_{k=1}^K f(\boldsymbol{x}_{\bar{S}}^{k},\boldsymbol{x}_S^*).
\end{equation}

In Kernel SHAP, $\boldsymbol{x}_{\bar{S}}^{k}, k=1,\ldots,K$ are sampled from the $\bar{S}$-part of 
the training data, *independently* of $\boldsymbol{x}_{S}$. This is motivated by using the 
training set as the empirical distribution of $\boldsymbol{x}_{\bar{S}}$, and assuming that 
$\boldsymbol{x}_{\bar{S}}$ is independent of $\boldsymbol{x}_S=\boldsymbol{x}_S^*$.
Due to the independence assumption, if the features in a given model are highly dependent, the 
Kernel SHAP method may give a completely wrong answer. This can be avoided by estimating the 
conditional distribution $p(\boldsymbol{x}_{\bar{S}}|\boldsymbol{x}_S=\boldsymbol{x}_S^*)$ directly 
and generating samples from this distribution. With this small change, the contributions and 
Shapley values may then be approximated as in the ordinary Kernel SHAP framework. @aas2019explaining 
propose three different approaches for estimating the conditional probabilities. The methods may 
also be combined, such that e.g. one method is used when conditioning on a small number of features, 
while another method is used otherwise.


## Multivariate Gaussian Distribution Approach

The first approach arises from the assumption that the feature vector $\boldsymbol{x}$ stems from a 
multivariate Gaussian distribution with some mean vector $\boldsymbol{\mu}$ and covariance matrix 
$\boldsymbol{\Sigma}$. Under this assumption, the conditional distribution 
$p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$ 
is also multivariate Gaussian  $\text{N}_{|\bar{\mathcal{S}}|}(\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}},\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}})$, 
with analytical expressions for the conditional mean vector $\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ 
and covariance matrix $\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$, see @aas2019explaining for details.
Hence, instead of sampling from the marginal empirical distribution of $\boldsymbol{x}_{\bar{\mathcal{S}}}$ 
approximated by the training data, we can sample from the Gaussian conditional distribution, which is fitted 
using the training data. Using the resulting samples 
$\boldsymbol{x}_{\bar{\mathcal{S}}}^k, k=1,\ldots,K$, the conditional expectations be approximated 
as in the Kernel SHAP. 


## Gaussian Copula Approach

If the features are far from multivariate Gaussian, an alternative approach is to instead represent 
the marginals by their empirical distributions, and model the dependence structure by a Gaussian 
copula. Assuming a Gaussian copula, we may convert the marginals of the training data to Gaussian 
features using their empirical distributions, and then fit a multivariate Gaussian distribution to these. 

To produce samples from the conditional distribution $p(\boldsymbol{x}_{\bar{\mathcal{S}}} |\boldsymbol{x}_{\mathcal{S}}=\boldsymbol{x}_{\mathcal{S}}^*)$, 
we convert the marginals of $\boldsymbol{x}_{\mathcal{S}}$ to Gaussians, sample from the conditional
Gaussian distribution as above, and convert the marginals of the samples back to the original 
distribution. Those samples are then used to approximate the sample from the resulting multivariate 
Gaussian conditional distribution. While other copulas may be used, the Gaussian copula has the 
benefit that we may use the analytical expressions for the conditionals 
$\boldsymbol{\mu}_{\bar{\mathcal{S}}|\mathcal{S}}$ and $\boldsymbol{\Sigma}_{\bar{\mathcal{S}}|\mathcal{S}}$. 
Finally, we may convert the marginals back to their original distribution, and use the resulting 
samples to approximate the conditional expectations as in the Kernel SHAP.  

## Empirical Conditional Distribution Approach

If both the dependence structure and the marginal distributions of $\boldsymbol{x}$ are very far 
from the Gaussian, neither of the two aforementioned methods will work very well. Few methods 
exists for the non-parametric estimation of conditional densities, and the classic kernel 
estimator (@rosenblatt1956) for non-parametric density estimation suffers greatly from the 
curse of dimensionality and does not provide a way to generate samples from the estimated 
distribution. For such situations, @aas2019explaining propose an empirical conditional approach 
to sample approximately from $p(\boldsymbol{x}_{\bar{\mathcal{S}}}|\boldsymbol{x}_{\mathcal{S}}^*)$. 
The idea is to compute weights $w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^i),\ i=1,...,n_{\text{train}}$ 
for all training instances based on their Mahalanobis distances (in the $S$ subset only) to the 
instance $\boldsymbol{x}^*$ to be explained. Instead of sampling from this weighted (conditional) 
empirical distribution, @aas2019explaining suggests a more efficient variant, using only the $K$ 
instances with the largest weights:

$$v_{\text{condKerSHAP}}(\mathcal{S}) = \frac{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^{[k]}) f(\boldsymbol{x}_{\bar{\mathcal{S}}}^{[k]},\boldsymbol{x}_{\mathcal{S}}^*)}{\sum_{k=1}^K w_{\mathcal{S}}(\boldsymbol{x}^*,\boldsymbol{x}^{[k]})},$$ 

The number of samples $K$ to be used in the approximate prediction can for instance be chosen such 
that the $K$ largest weights accounts for a fraction $\eta$, for example $0.9$, of the total weight. 
If $K$ exceeds a certain limit, for instance $5,000$, it might be set to that limit. A bandwidth
parameter $\sigma$ used to scale the weights, must also be specified. This choice may be viewed as 
a bias-variance trade-off. A small $\sigma$ puts most of the weight to a few of the closest 
training observations and thereby gives low bias, but high variance. When $\sigma \rightarrow \infty$, 
this method converges to the original Kernel SHAP assuming feature independence. Typically, when 
the features are highly dependent, a small $\sigma$ is typically needed such that the bias does 
not dominate. @aas2019explaining show that a proper criterion for selecting $\sigma$ is a 
small-sample-size corrected version of the AIC known as AICc. As calculation of it is 
computationally intensive, an approximate version of the selection criterion is also suggested. 
Details on this is found in @aas2019explaining.

<a id="ex"></a>

<br>

# Examples

`shapr` supports computation of Shapley values with any predictive model which takes a set of 
numeric features and produces a numeric outcome. 

The following example shows how a simple `xgboost` model is trained using the Boston Housing Data, 
and how `shapr` can be used to explain the individual predictions. Note that empirical conditional 
distribution approach is the default (i.e. `approach = "empirical"`), and that the Gaussian and 
Gaussian copula approaches can be used instead by setting the argument `approach` to either 
`"gaussian"` or `"copula"`.

```{r, warning=FALSE}
library(xgboost)
library(shapr)

data("Boston", package = "MASS")

x_var <- c("lstat", "rm", "dis", "indus")
y_var <- "medv"

x_train <- as.matrix(Boston[-(1:6), x_var])
y_train <- Boston[-(1:6), y_var]
x_test <- as.matrix(Boston[1:6, x_var])

# Fitting a basic xgboost model to the training data
model <- xgboost(
  data = x_train,
  label = y_train,
  nround = 20,
  verbose = FALSE
)

# Prepare the data for explanation
explainer <- shapr(x_train, model)

# Specifying the phi_0, i.e. the expected prediction without any features
pred_zero <- mean(y_train)

# Computing the actual Shapley values with kernelSHAP accounting for feature 
# dependence using the empirical (conditional) distribution approach with 
# bandwidth parameter sigma = 0.1 (default)
explanation <- explain(
  x_test,
  approach = "empirical",
  explainer = explainer,
  prediction_zero = pred_zero
)

# Printing the Shapley values for the test data
explanation$dt

# Plot the resulting explanations for observations 1 and 6
plot(explanation, plot_phi0 = FALSE, index_x_test = c(1,6))
``` 

The Gaussian approach is used as follows:

```{r}
# Use the Gaussian approach
explanation_gaussian <- explain(
  x_test,
  approach = "gaussian",
  explainer = explainer,
  prediction_zero = pred_zero
)

# Plot the resulting explanations for observations 1 and 6
plot(explanation_gaussian, plot_phi0 = FALSE, index_x_test = c(1,6))
```

The Gaussian copula approach is used as follows:

```{r}
# Use the Gaussian copula approach
explanation_copula <- explain(
  x_test,
  approach = "copula",
  explainer = explainer,
  prediction_zero = pred_zero
)

# Plot the resulting explanations for observations 1 and 6, excluding 
# the no-covariate effect
plot(explanation_copula, plot_phi0 = FALSE, index_x_test = c(1,6))
```

## Main arguments in `shapr`
When using `shapr`, the default behavior is to use all feature combinations in the Shapley 
formula. Kernel SHAPs sampling based approach may be used by specifying `n_combinations`, which 
is the number of feature combinations to sample. If not specified, the exact method is used. 
The computation time grows approximately exponentially with the number of samples. The training data 
and the model whose predictions we wish to explain must be provided through 
the arguments \code{x} and \code{model}. Note that `x` must be a `data.frame` or a `matrix`, and all
elements must be finite numerical values. Currently we do not support categorical features or 
missing values.
 

## Main arguments in `explain`

The test data given by `x`, whose predicted values we wish to explain, must be provided. Note that 
`x` must be a `data.frame` or a `matrix,` where all elements are finite numerical values.
One must also provide the object returned by `shapr` through the argument `explainer`. 
The approach to use when computing the Shapley values must also be specified through the argument 
`approach`, either equal to `"empirical"`, `"gaussian"` or `"copula"`, or a vector 
of them. If a vector, a combined approach is used, and element `i` indicates the approach to 
use when conditioning on `i` variables. Similarly, `type` is only applicable if the 
`approach = "empirical"`, and indicates the type of empirical approach to use. It is either 
`"independence"`, `"fixed_sigma"` or `"gaussian"`.

When computing the kernel SHAP values by `explain`, the maximum number of samples to use in the 
Monte Carlo integration for every conditional expectation is controlled by the argument `n_samples` 
(default equals `1000`). The computation time grows approximately linear with this number. As 
seen in the code above, the argument `prediction_zero` for the prediction value of the unseen 
data is required. We recommend setting this equal to the mean of the response, but other values, 
like the mean prediction of a large test data set is also a possibility.
If the empirical method is used, specific settings for that approach, like a fixed $\sigma$ value 
of vector can be specified through the argument `sigma_vec`. If the `approach = "gaussian"`,
we may specify the mean vector and covariance matrix of the data generating distribution by the 
arguments `mu` and `cov_mat`. If not specified, they are estimated from the training data.  

<a id="compare"></a>

<br>

# Comparison to Lundberg & Lee's implementation

As mentioned above, the original (independence assuming) Kernel SHAP implementation can be 
approximated by setting a large $\sigma$ value using our empirical approach. If we specify that the 
distances to *all* training observations should be used (i.e. setting `approach = "empirical"` and
`w_threshold = 1` when using `explain`, we can approximate the original method arbitrarily 
well by increasing $\sigma$. For completeness of the `shapr`, we have also implemented 
a version of the original method, which samples training observations independently with respect to 
their distances to test observations (i.e. without the large-$\sigma$ approximation). This method is 
available by using `approach = "empirical"` and `type = "independence"` in `explain`. 

We have compared the results using these two variants with the original implementation of 
@lundberg2017unified, available through the Python library [`shap`](https://github.com/slundberg/shap).
As above, we used the Boston housing 
data, trained via `xgboost`. We specify that *all* training observations should be used when 
explaining all of the 6 test observations. To run the individual explanation method in the `shap` 
Python library we use the `reticulate` `R`-package, allowing Python code to run within `R`. 
As this requires installation of Python package, the comparison code and results is not included 
in this vignette, but can be found [here](https://github.com/NorskRegnesentral/shapr/blob/master/inst/scripts/compare_lundberg.R).
As indicated by the (commented out) results in the file above both methods in our
`R`-package give (up to numerical approximation error) identical results to the original 
implementation in the Python  `shap` library. 

<br>

# References

